### 2024-02-16 manuscript/manuscript.Rmd
Intro paragraph ideation coming

---
title: "Deep Learning Models for Limnological Time Series Forecasting"
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mlapeyro@berkeley.edu
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu (corresponding author)
abstract: |
  1. 
  2. 
  3. 
  4.

keywords:
  - Artificial Intelligence
  - Forecasting
  - Machine Learning
  - Time Series
  - Limnology

bibliography: references.bib
nocite: |

header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \usepackage{bbm}
  - \usepackage{setspace}
  - \linenumbers
  - \doublespacing

output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

# Introduction

The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is significant to natural resource management. Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to circumvent scenarios with far-reaching negative impacts. Yet, forecasting these variables is challenging because they are influenced by many processes that are hard to predict and characterize. Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}. A shortcoming of these studies which support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of sites in different geographic regions. This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.

The time series that are used in this study are taken from the National Ecological Observatory Network's Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to the publicly accessible data \cite{Quinn}. A common theme across the challenge is that a historical means (or climatology) model tends to be a top performing model \cite{Dietze, Quinn}. Thus, a primary consideration in this paper is the comparison of the machine learning models to the climatology model. If machine learning models can consistently outperform the climatology model, then machine learning is clearly a promising method class for limonological time series forecasting.

** Talk about NEON forecasting challenge and how this offers an unprecedented opportunity to test machine learning models. Typically ML models have been applied to a handful of freshwater systems; the forecasting challenge offers 34 sites across North America and Puerto Rico as a testing ground. Historically, researchers have favored using process-based models to make forecasts, but building mechanistic models for 34 sites is very challenging. Thus, leveraging non-mechanistic machine learning models offer some practical advantages. Machine learning models have been applied to limnological time series forecasting at varying levels of success. Generally, the machine learning models tended to be more basic, like RNN's and LSTM's, in the interim, new models have been developed that are said to outperform these earlier ML models. This study makes a comparison of some the newly touted state of the art machine learning models on the aquatics time series taken from NEON. From a current literature search, no previous paper has made as comprehensive of a comparison of machine learning models on the variables of water temperature, chlorophyll concentration and dissolved oxygen.

# Materials and Methods

## Preprocessing

## Null Models

## Machine Learning Models

# Results

# Discussion

# Acknowledgements

# Conflicts of Interest {-}

The authors declare there are no conflicts of interest.

# Authors' Contributions {-}

Marcus Lapeyrolerie and Carl Boettiger developed the code and wrote the manuscript.

# Data Availability {-}

\newpage

# References

### 2024-02-16 manuscript/manuscript.Rmd
Intro paragraph ideation coming redux

@@
 A common theme across the challenge is that a historical means (or climatology) model tends to be a top performing model \cite{Dietze, Quinn}.
+In a model comparison for the phenology challenge, \cite{Dietze} found that the climatology outperformed all but one of the submitted models, and the best performing forecast only marginally outperformed the climatology model.
 Thus, a primary consideration in this paper is the comparison of the machine learning models to the climatology model.
-If machine learning models can consistently outperform the climatology model, then machine learning is clearly a promising method class for limonological time series forecasting.
+If deep learning models can consistently outperform the climatology model, then deep learning is clearly a promising method class for limonological time series forecasting.
 
-** Talk about NEON forecasting challenge and how this offers an unprecedented opportunity to test machine learning models.
-Typically ML models have been applied to a handful of freshwater systems; the forecasting challenge offers 34 sites across North America and Puerto Rico as a testing ground.
-Historically, researchers have favored using process-based models to make forecasts, but building mechanistic models for 34 sites is very challenging.
-Thus, leveraging non-mechanistic machine learning models offer some practical advantages.
-Machine learning models have been applied to limnological time series forecasting at varying levels of success.
-Generally, the machine learning models tended to be more basic, like RNN's and LSTM's, in the interim, new models have been developed that are said to outperform these earlier ML models.
-This study makes a comparison of some the newly touted state of the art machine learning models on the aquatics time series taken from NEON.
-From a current literature search, no previous paper has made as comprehensive of a comparison of machine learning models on the variables of water temperature, chlorophyll concentration and dissolved oxygen.
+Talk about Darts, and how there have been newly developed deep learning models that offer advantages over the more vanilla ones that are commonly used.
+The advantage of taking the machine learning approach is that it allows the ability to easily deploy models on systems without having a mechanistic understanding of them.
+This is especially advantageous in the current era where sensors can be deployed at a much faster rate than we can build accurate water quality models for every existent body of water.
+Like Sci-kit Learn or Tidy Models, Darts allows users to easily instantiate, fit and predict with these state of the art models without requiring users to create these models from scratch.
 

### 2024-02-17 manuscript/manuscript.Rmd
intro rough draft done

@@
 ---
-title: "Deep Learning Models for Limnological Time Series Forecasting"
+title: "A Comparison of Machine Learning Models for Limnological Time Series Forecasting"
 authors:
@@
 
-The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is significant to natural resource management.
+The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is important to natural resource management.
 Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to circumvent scenarios with far-reaching negative impacts.
@@
 Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}.
-A shortcoming of these studies which support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of sites in different geographic regions.
+A shortcoming of these studies in support of machine learning is that they tend to focus on a handful of sites that are not necessarily representative of bodies of water across larger geographic regions.
 This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.
 
-The time series that are used in this study are taken from the National Ecological Observatory Network's Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to the publicly accessible data \cite{Quinn}.
-A common theme across the challenge is that a historical means (or climatology) model tends to be a top performing model \cite{Dietze, Quinn}.
-In a model comparison for the phenology challenge, \cite{Dietze} found that the climatology outperformed all but one of the submitted models, and the best performing forecast only marginally outperformed the climatology model.
-Thus, a primary consideration in this paper is the comparison of the machine learning models to the climatology model.
-If deep learning models can consistently outperform the climatology model, then deep learning is clearly a promising method class for limonological time series forecasting.
+The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{Quinn}.
+A common finding across the challenge is that a historical means model (also referred to as the climatology model) tends to generate top scoring forecasts \cite{Dietze, Quinn}.
+For instance, in a model comparison that examined the forecasts for the phenology data from the Ecological Forecasting Challenge, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
+Thus, a primary consideration in this paper is the comparing the performance of the machine learning models to the climatology model.
 
-Talk about Darts, and how there have been newly developed deep learning models that offer advantages over the more vanilla ones that are commonly used.
-The advantage of taking the machine learning approach is that it allows the ability to easily deploy models on systems without having a mechanistic understanding of them.
-This is especially advantageous in the current era where sensors can be deployed at a much faster rate than we can build accurate water quality models for every existent body of water.
-Like Sci-kit Learn or Tidy Models, Darts allows users to easily instantiate, fit and predict with these state of the art models without requiring users to create these models from scratch.
+There is a large variety of neural network architectures that are used for time series forecasting.
+Past applications of machine learning to limnological time series forecasting have used architectures like the long-short-term-memory (LSTM) network which was published in 1997 \cite{lstm}.
+Over the nearly 30 years since LSTM's were introduced, there have been a plethora of other neural network architectures that have been developped for time series forecasting, many of which were designed to avoid the shortcomings inherent to earlier architectures like LSTM's \cite{some_ml_papers}.
+This paper will compare 8 machine learning models including LSTM's to more recently developed ML methods for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.
 
@@
 
+Talk generally about how forecasts were produced.
+we looked at 12 different time intervals over the span BLANK.
+8 machine learning models, and we compared these against 2 null models, a naive persistence model and a climatology model.
+For the climatology model, we compared the probabilistic forecasts using the CRPS metric.
+For the naive persistence model, we compared the point forecast, using the median of the probabilistic forecast generated by the machine learning models.
+
 ## Preprocessing
+
+Talk about how preprocessing was performed using daily mean data whenever possible.
 
@@
 
+Talk briefly about how you defined the naive persistence and climatology model.
+
 ## Machine Learning Models
 
+Introduce darts, talk about the range of machine learning models that were included.
+The models generate probabilistic forecasts, this is because the model attempts to do quantile regression.
+The model will then sample from the distribution specified by the ML model outputs and this will serve as the forecast.
+
 # Results
+
+Plots. Talk about plots.
 

### 2024-02-24 manuscript/manuscript.Rmd
getting there

@@
 The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is important to natural resource management.
-Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to circumvent scenarios with far-reaching negative impacts.
-Yet, forecasting these variables is challenging because they are influenced by many processes that are hard to predict and characterize.
+Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to prepare for scenarios with far-reaching negative impacts.
 Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}.
-A shortcoming of these studies in support of machine learning is that they tend to focus on a handful of sites that are not necessarily representative of bodies of water across larger geographic regions.
+A shortcoming of the limnological studies that support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of water systems across a continental scale.
 This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.
@@
 The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{Quinn}.
-A common finding across the challenge is that a historical means model (also referred to as the climatology model) tends to generate top scoring forecasts \cite{Dietze, Quinn}.
+A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}.
 For instance, in a model comparison that examined the forecasts for the phenology data from the Ecological Forecasting Challenge, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
@@
 
-There is a large variety of neural network architectures that are used for time series forecasting.
-Past applications of machine learning to limnological time series forecasting have used architectures like the long-short-term-memory (LSTM) network which was published in 1997 \cite{lstm}.
-Over the nearly 30 years since LSTM's were introduced, there have been a plethora of other neural network architectures that have been developped for time series forecasting, many of which were designed to avoid the shortcomings inherent to earlier architectures like LSTM's \cite{some_ml_papers}.
-This paper will compare 8 machine learning models including LSTM's to more recently developed ML methods for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.
+There are many neural network architectures that have not been explored for ecological time series forecasting.
+Past applications of machine learning to limnological time series forecasting have typically used the long-short-term-memory (LSTM) neural network which was first introduced in 1997 \cite{lstm}.
+Over the nearly 30 years since LSTM's were introduced, there has been much development in the application of machine learning to time series forecasting, but recent studies examining machine learning for ecological time series forecasting have not explored more contemporary neural network architectures.
+This paper will compare 8 machine learning algorithms including LSTM's to ML methods that have been recently developed for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.
 
@@
 
-Talk generally about how forecasts were produced.
-we looked at 12 different time intervals over the span BLANK.
-8 machine learning models, and we compared these against 2 null models, a naive persistence model and a climatology model.
-For the climatology model, we compared the probabilistic forecasts using the CRPS metric.
-For the naive persistence model, we compared the point forecast, using the median of the probabilistic forecast generated by the machine learning models.
+We used ML models to generate water temperature, dissolved oxygen and chlorophyll-A concentration forecasts for 34 freshwater sites across North America.
+These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes designated by NEON.
+We trained 8 different ML models for every site and target variable available under the Aquatics theme of the NEON Ecological Forecasting Challenge.
+For the models that accept past covariates (**list these models**), we inputted as covariates the other target variables recorded at that site as well as air temperature.
+For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate.
+The collected time series were split into a training and a validation set, whereby the models were trained on time series from 2018 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
 
-## Preprocessing
+We compared the ML models to a climatology and naive persistence null model.
+The climatology model generates forecasts by finding the daily mean and standard deviation of the target time series, and then drawing samples from a Gaussian Distribution with these parameters.
+The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast window.
+We evaluate the continuous rank probability scores for the climatology model, and the absolute error for the Naive Persistence model.
 
-Talk about how preprocessing was performed using daily mean data whenever possible.
+The imputation of missing data proved to be critical to the performance of the machine learning models.
+After experimenting with data filling methods that resulted in poor model performance, we settled on a imputation method inspired by the climatology model.
+Missing data is estimated using the daily historical mean when this statistic is available; and, for the case when there are no data collected for a day of the year, either the monthly, seasonal or global median was used with preference given to the median available from the shortest timescale.
 
-## Null Models
-
-Talk briefly about how you defined the naive persistence and climatology model.
+We provide fully reproducible code for fitting, scoring and visualizing the forecasts.
+All the machine learning forecasts were created using the `darts` python library.
+`darts` is similar to other libraries `scikit-learn` in Python or `tidymodels` in R that allow the user to use complex time series forecasting models without having to implement them.
+Otherwise, we rely on the `CRPS` library in python to efficiently compute the continuous ranked probability scores (CRPS).
+When comparing the probabilistic forecasts, we express the CRPS score in error-orientation -- i.e., a lower CRPS indicates a smaller margin of error.
 
@@
 
-Introduce darts, talk about the range of machine learning models that were included.
-The models generate probabilistic forecasts, this is because the model attempts to do quantile regression.
-The model will then sample from the distribution specified by the ML model outputs and this will serve as the forecast.
+All the machine learning algorithms that are used in this study are based on neural network architectures.
+Neural networks have the property of being universal function approximators, so theoretically, it is possible that all of these models could exactly approximate the data generating process \cite{}.
+Yet, it is often true that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock, makridakis}.
+This performance gap can be due to a variety of reasons including overfitting and insufficient hyperparameter tuning \cite{adcock}.
+The discrepancy between theory and practice in function approximation with neural networks motivates research on how machine learning models perform in specific domains.
+
+The 8 machine learning models that we compare take a variety of approaches with the design of their neural networks.
+We will not go into detail on how these different models work, but for those interested to learn more, we list the models and their references in table \ref{}.
+For readers who do not specialize in ML, an important concept to understand is that the ML models learn directly from the data and are not instilled with domain knowledge.
+This non-mechanisitic basis is at once very powerful as it does not restrict the models with misleading assumptions but is simultaneously limiting as these models are not readily interpretable and generally require more data than mechanistic analogs \cite{}.
+For the management of critical resources like water, this lack of interpretability could be a deterrent to the use of ML methods, and there may not be enough data to accomodate ML approaches \cite{}.
+
+The ML models are configured to generate probabilistic forecasts.
+The neural network architectures that are used in this study are deterministic, so they are not able to directly represent probability distributions.
+We work around this limitation by performing quantile regression whereby the neural networks are trained to output quantiles at each time step in the forecast window.
+Forecasts are then generated by drawing samples according to these quantiles.
 
@@
 # Discussion
+
+Over the last decade, ML has gained significant prominence after a spate of high profile successes like beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}.
+During this time, researchers have developed many ML models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}.
+For this study, we selected 8 ML models that include s
+
+Yet, these claims of improvement tend to have major limitations, notably that the studies examine a limited amount of time series and they do not compare to naive benchmarks \cite{makridakis}.
 

### 2024-02-27 manuscript/manuscript.Rmd
:pen:

@@
 Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}.
-A shortcoming of the limnological studies that support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of water systems across a continental scale.
+A shortcoming of the limnological studies that support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of water systems on a broad geographic scale.
 This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.
@@
 A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}.
-For instance, in a model comparison that examined the forecasts for the phenology data from the Ecological Forecasting Challenge, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
-Thus, a primary consideration in this paper is the comparing the performance of the machine learning models to the climatology model.
+For instance, in a model comparison that examined the forecasts for phenology time series, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
+Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the climatology model.
 
@@
 Past applications of machine learning to limnological time series forecasting have typically used the long-short-term-memory (LSTM) neural network which was first introduced in 1997 \cite{lstm}.
-Over the nearly 30 years since LSTM's were introduced, there has been much development in the application of machine learning to time series forecasting, but recent studies examining machine learning for ecological time series forecasting have not explored more contemporary neural network architectures.
+Over the nearly 30 years since LSTM's were introduced, there has been much development in the subfield of machine learning for time series forecasting, but recent studies examining machine learning for ecological time series forecasting have not explored more contemporary neural network architectures.
 This paper will compare 8 machine learning algorithms including LSTM's to ML methods that have been recently developed for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.
@@
 
-We used ML models to generate water temperature, dissolved oxygen and chlorophyll-A concentration forecasts for 34 freshwater sites across North America.
-These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes designated by NEON.
-We trained 8 different ML models for every site and target variable available under the Aquatics theme of the NEON Ecological Forecasting Challenge.
-For the models that accept past covariates (**list these models**), we inputted as covariates the other target variables recorded at that site as well as air temperature.
+We employed ML models to generate water temperature, dissolved oxygen and chlorophyll-A concentration forecasts for 34 freshwater sites across North America.
+These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON.
+8 different ML models were trained for every site and target variable available in NEON Ecological Forecasting Challenge Aquatics theme.
+For the models that accept past covariates (**list these models**), we used the other target variables recorded at that site as well as air temperature as covariates.
 For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate.
-The collected time series were split into a training and a validation set, whereby the models were trained on time series from 2018 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
+The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2018 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
 
@@
 The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast window.
-We evaluate the continuous rank probability scores for the climatology model, and the absolute error for the Naive Persistence model.
 
@@
 We provide fully reproducible code for fitting, scoring and visualizing the forecasts.
-All the machine learning forecasts were created using the `darts` python library.
-`darts` is similar to other libraries `scikit-learn` in Python or `tidymodels` in R that allow the user to use complex time series forecasting models without having to implement them.
-Otherwise, we rely on the `CRPS` library in python to efficiently compute the continuous ranked probability scores (CRPS).
+All the machine learning forecasts were generated using the `darts` python library.
+`darts` is similar to other libraries `scikit-learn` in Python or `tidymodels` in R that allow users to easily employ time series forecasting models without having to implement them.
+For model comparison, we rely on the `CRPS` library in Python to efficiently compute the continuous ranked probability scores (CRPS).
 When comparing the probabilistic forecasts, we express the CRPS score in error-orientation -- i.e., a lower CRPS indicates a smaller margin of error.
@@
 # Results
-
-Plots. Talk about plots.
 

### 2024-02-27 train_tune.py
Forgot to add train/validate distinction here

@@ -86,10 +86,16 @@ if __name__ == "__main__":
     covariates_list.remove(args.target)
     if args.nocovs:
         covariates_list = None
-    
-    data_preprocessor = TimeSeriesPreprocessor(input_csv_name = "targets.csv.gz",
-                                               load_dir_name = "preprocessed_timeseries/")
-    data_preprocessor.load(args.site)
+
+    # Loading data preprocessors for training and validation
+    data_preprocessors = []
+    for suffix in ['train', 'validate']:
+        preprocessor = TimeSeriesPreprocessor(
+            input_csv_name = "targets.csv.gz",
+            load_dir_name = f"preprocessed_{suffix}/",
+        )
+        preprocessor.load(args.site)
+        data_preprocessors.append(preprocessor)
     
     # Handling csv names and directories for the final forecast
     if not os.path.exists(f"forecasts/{args.site}/{args.target}/"):
@@ -103,17 +109,20 @@ if __name__ == "__main__":
     # Instantiating the model
     extras = {"epochs": args.epochs,
               "verbose": args.verbose,}
-    forecaster = BaseForecaster(model=args.model,
-                        target_variable=args.target,
-                        data_preprocessor=data_preprocessor,
-                        covariates_names=covariates_list,
-                        output_csv_name=f"{output_csv_name}.csv",
-                        validation_split_date=args.date,
-                        model_hyperparameters=hyperparams_dict["model_hyperparameters"],
-                        model_likelihood=model_likelihood,
-                        site_id=args.site,
-                        num_trials=args.num_trials,
-                        **extras)
+    forecaster = BaseForecaster(
+        model=args.model,
+        target_variable=args.target,
+        train_preprocessor=preprocessors[0],
+        validate_preprocessor=preprocessors[0],
+        covariates_names=covariates_list,
+        output_csv_name=f"{output_csv_name}.csv",
+        validation_split_date=args.date,
+        model_hyperparameters=hyperparams_dict["model_hyperparameters"],
+        model_likelihood=model_likelihood,
+        site_id=args.site,
+        num_trials=args.num_trials,
+        **extras,
+    )
     
     # 
     if args.tune:

### 2024-04-13 jobs/preprocess_timeseries.py
Maybe running now

@@ -69,4 +69,4 @@ if __name__=="__main__":
     data_preprocessor.save()
 
     
-    print('\n' + f"Runtime: {(time.time() - start_)/60:.2f} minutes")
+    print('\n' + f"Runtime: {(time.time() - start)/60:.2f} minutes" + '\n')

### 2024-04-13 save_plots.py

@@ -49,7 +49,7 @@ try:
     )
     s3_dict = {'client': s3_client, 'bucket': args.bucket}
 except:
-    s3_client = None
+    s3_dict = {'client': None, 'bucket': None}
 
 # Lists and dictionaries to store dataframes and 
 best_performers_dfs = {}

### 2024-04-13 train_tune.py

@@ -14,6 +14,9 @@ import os
 import copy
 import json
 import yaml
+import numpy as np
+from datetime import datetime, timedelta
+
 start = time.time()
 
 
@@ -25,10 +28,6 @@ parser.add_argument("--target", default="oxygen", type=str,
                     "[oxygen, temperature, chla].")
 parser.add_argument("--site", default="BARC", type=str,
                     help="Denotes which site to use.")
-parser.add_argument("--date", default="2022-07-19", type=str,
-                    help="Flags for the validation split date, "+\
-                    "n.b. that this should align with last date " +\
-                    "of the preprocessed time series.")
 parser.add_argument("--epochs", default=200, type=int, 
                     help="The number of epochs to train a model for.")
 parser.add_argument("--nocovs", default=False, action="store_true",
@@ -70,14 +69,14 @@ if __name__ == "__main__":
         s3_client, s3_dict = None, None
         
     # Selecting the device
-    try:
-        os.environ["CUDA_VISIBLE_DEVICES"] = f"{args.device}"
-    except:
-        continue
+    os.environ["CUDA_VISIBLE_DEVICES"] = f"{args.device}"
 
     # Accessing the validation split date from targets csv
     targets = pd.read_csv("aquatics-targets.csv.gz")
-    validation_split_date = np.sort(targets['datetime'].unique())[-1]
+    most_recent_date_str = np.sort(targets['datetime'].unique())[-1]
+    most_recent_date = datetime.strptime(most_recent_date_str, '%Y-%m-%d')
+    one_year_before = most_recent_date - timedelta(days=365)
+    validation_split_date = one_year_before.strftime('%Y-%m-%d')
     
     # Loading hyperparameters
     hyperparams_loc = f"hyperparameters/train/{args.target}/{args.model}"

### 2024-04-16 manuscript/manuscript.Rmd
Plotting coming together

@@
 ---
-title: "A Comparison of Machine Learning Models for Limnological Time Series Forecasting"
+title: "A Comparison of Machine Learning Models for Water Quality Forecasting"
 authors:
@@
   - \usepackage{setspace}
+  - \usepackage{graphicx}
+  - \usepackage{subcaption}
   - \linenumbers
@@
 
+\begin{figure}[!htbp]
+  \centering
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/global_historical_oxygen.png}
+    \caption{}
+  \end{subfigure}
+
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/global_naive_oxygen.png}
+    \caption{}
+  \end{subfigure}
+
+  \caption{Comparison of Historical and Naive Oxygen}
+\end{figure}
+
+\begin{figure}[!htbp]
+  \centering
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/global_historical_temperature.png}
+    \caption{}
+  \end{subfigure}
+
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/global_naive_temperature.png}
+    \caption{}
+  \end{subfigure}
+
+  \caption{Comparison of Historical and Naive temperature}
+\end{figure}
+
+\begin{figure}[!htbp]
+  \centering
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/global_historical_chla.png}
+    \caption{}
+  \end{subfigure}
+
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/global_naive_chla.png}
+    \caption{}
+  \end{subfigure}
+
+  \caption{Comparison of Historical and Naive chla}
+\end{figure}
+
+\begin{figure}[!htbp]
+  \centering
+
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_historical_oxygen.png}
+    \caption{}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_historical_temperature.png}
+    \caption{}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_historical_chla.png}
+    \caption{}
+  \end{subfigure}
+
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_naive_oxygen.png}
+    \caption{}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_naive_temperature.png}
+    \caption{}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_naive_chla.png}
+    \caption{}
+  \end{subfigure}
+
+  \caption{Comparison of Historical and Naive parameters intra}
+\end{figure}
+
+\begin{figure}[!htbp]
+  \centering
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/global_loc_historical_oxygen.png}
+    \caption{}
+  \end{subfigure}
+
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/date_type_global_historical_oxygen.png}
+    \caption{}
+  \end{subfigure}
+
+  \caption{Comparison of Historical and Naive Oxygen}
+\end{figure}
+
 # Discussion

### 2024-04-16 plotting_utils.py

@@ -737,14 +737,17 @@ def plot_global_percentages(df_, historical=True, png_name=None):
     plt.grid(False)
     plt.axhline(y=0, color='black', linestyle='dashed', linewidth=1)
     if historical:
-        plt.ylabel("CRPSS")
+        plt.ylabel("CRPSS", fontsize=20)
     else:
-        plt.ylabel("RMSE-SS")
+        plt.ylabel("RMSE-SS", fontsize=20)
     ax = plt.gca()
     ax.spines["left"].set_visible(True)
     ax.spines["bottom"].set_visible(True)
-    plt.xticks(rotation=30)
+    plt.xlabel('model', fontsize=20)
+    plt.xticks(rotation=30, fontsize=18)
+    plt.yticks(fontsize=18)
     plt.legend(labels=[])
+    plt.tight_layout()
 
     # Saving the plot if desired
     if png_name:
@@ -987,7 +990,7 @@ def plot_crps_over_time_agg(intra_df, historical=True, png_name=None):
         y='50th_percentile',
         legend=False,
         color='#1f77b4',
-        linewidth=2,
+        linewidth=6,
     )
     
     # Customize plot appearance
@@ -996,12 +999,14 @@ def plot_crps_over_time_agg(intra_df, historical=True, png_name=None):
     ax.spines["left"].set_visible(True)
     ax.spines["bottom"].set_visible(True)
     plt.axhline(y=0, color='black', linestyle='dashed', linewidth=1)
-    plt.legend(labels=['ML Model Aggregate'], loc='upper right')
     if metric == 'ae':
-        plt.ylabel("AbsErr-SS")
+        plt.ylabel("AbsErr-SS", fontsize=40)
     elif metric == 'crps':
-        plt.ylabel("CRPSS")
-    plt.show()
+        plt.ylabel("CRPSS", fontsize=40)
+    plt.xlabel("t", fontsize=40)
+    plt.xticks(fontsize=38)
+    plt.yticks(fontsize=38)
+    plt.tight_layout()
 
     # Saving the plot if desired
     if png_name:

### 2024-04-18 manuscript/manuscript.Rmd
Caught some minor bugs

@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/global_historical_oxygen.png}
-    \caption{}
+    \includegraphics[width=\textwidth]{../plots/type_local_historical_oxygen.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/global_naive_oxygen.png}
-    \caption{}
+    \includegraphics[width=\textwidth]{../plots/type_local_naive_oxygen.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/global_historical_temperature.png}
-    \caption{}
+    \includegraphics[width=\textwidth]{../plots/type_local_historical_temperature.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/global_naive_temperature.png}
-    \caption{}
+    \includegraphics[width=\textwidth]{../plots/type_local_naive_temperature.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/global_historical_chla.png}
-    \caption{}
+    \includegraphics[width=\textwidth]{../plots/type_local_historical_chla.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/global_naive_chla.png}
-    \caption{}
+    \includegraphics[width=\textwidth]{../plots/type_local_naive_chla.png}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{../plots/intra_historical_oxygen.png}
-    \caption{}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{../plots/intra_historical_temperature.png}
-    \caption{}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{../plots/intra_historical_chla.png}
-    \caption{}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{../plots/intra_naive_oxygen.png}
-    \caption{}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{../plots/intra_naive_temperature.png}
-    \caption{}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{../plots/intra_naive_chla.png}
-    \caption{}
   \end{subfigure}
 
-  \caption{Comparison of Historical and Naive parameters intra}
+  \caption{Oxygen, Temperature, Chlorophyll Intrawindow plots}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{../plots/global_loc_historical_oxygen.png}
-    \caption{}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{../plots/date_type_global_historical_oxygen.png}
-    \caption{}
   \end{subfigure}
 
-  \caption{Comparison of Historical and Naive Oxygen}
+  \caption{Global performance on oxygen}
 \end{figure}

### 2024-04-18 train_tune.py

@@ -70,9 +70,10 @@ if __name__ == "__main__":
         
     # Selecting the device
     os.environ["CUDA_VISIBLE_DEVICES"] = f"{args.device}"
+    csv_name = "aquatics-targets.csv.gz"
 
     # Accessing the validation split date from targets csv
-    targets = pd.read_csv("aquatics-targets.csv.gz")
+    targets = pd.read_csv(csv_name)
     most_recent_date_str = np.sort(targets['datetime'].unique())[-1]
     most_recent_date = datetime.strptime(most_recent_date_str, '%Y-%m-%d')
     one_year_before = most_recent_date - timedelta(days=365)
@@ -98,7 +99,7 @@ if __name__ == "__main__":
     preprocessors = []
     for suffix in ['train', 'validate']:
         preprocessor = TimeSeriesPreprocessor(
-            input_csv_name = "targets.csv.gz",
+            input_csv_name = csv_name,
             load_dir_name = f"preprocessed_{suffix}/",
             s3_dict=s3_dict,
         )

### 2024-04-18 utils.py

@@ -339,7 +339,7 @@ def season_doy_range(year, month, day):
 
 class TimeSeriesPreprocessor():
     def __init__(self,
-                 input_csv_name = "targets.csv.gz",
+                 input_csv_name = "aquatics-targets.csv.gz",
                  s3_dict: Optional[dict] = {'client': None, 'bucket': None},
                  load_dir_name: Optional[str] = "preprocessed_timeseries/",
                  datetime_column_name: Optional[str] = "datetime",
@@ -617,7 +617,7 @@ class BaseForecaster():
                  seed: Optional[int] = 0,
                  verbose: Optional[bool] = False,
                  log_tensorboard: Optional[bool] = False,
-                 targets_csv: Optional[str] = "targets.csv.gz",
+                 targets_csv: Optional[str] = "aquatics-targets.csv.gz",
                  s3_dict: Optional[dict] = {'client': None, 'bucket': None}
                  ):
         self.model_ = {"BlockRNN": BlockRNNModel, 
@@ -882,7 +882,6 @@ class NaiveEnsembleForecaster():
                  target_variable: Optional[str] = None,
                  date: Optional[str] = None,
                  output_name: Optional[str] = "default",
-                 targets_csv: Optional[str] = "targets.csv.gz",
                  s3_dict: Optional[dict] = {'client': None, 'bucket': None}
                  ):
         self.model_list = model_list
@@ -890,7 +889,6 @@ class NaiveEnsembleForecaster():
         self.target_variable = target_variable
         self.date = date
         self.output_name = output_name
-        self.targets_df = pd.read_csv(targets_csv)
         self.s3_dict = s3_dict
     
         if not self.s3_dict['client']:

### 2024-04-19 plotting_utils.py
AutoTheta now running

@@ -841,13 +841,15 @@ def plot_site_type_percentages_bymodel(df_, metadata_df, historical=True, png_na
     plt.grid(False)
     plt.axhline(y=0, color='black', linestyle='dashed', linewidth=1)
     if historical:
-        plt.ylabel("CRPSS")
+        plt.ylabel("CRPSS", fontsize=20)
     else:
-        plt.ylabel("RMSE-SS")
+        plt.ylabel("RMSE-SS", fontsize=20)
     ax = plt.gca()
     ax.spines["left"].set_visible(True)
     ax.spines["bottom"].set_visible(True)
-    plt.xticks(rotation=30)
+    plt.xlabel("model", fontsize=20)
+    plt.xticks(rotation=30, fontsize=18)
+    plt.yticks(fontsize=18)
     legend_handles = [Patch(facecolor=color, edgecolor='black') for color in color_dict.values()]
     legend_labels = list(color_dict.keys())
     plt.legend(legend_handles, legend_labels, title='Site Type', loc='lower right')

### 2024-04-19 predict_stats.py

@@ -1,0 +1,54 @@
+from stats_utils import AutoThetaForecaster
+from utils import (
+    TimeSeriesPreprocessor,
+    establish_s3_connection,
+    read_and_pivot_csv,
+)
+import time
+import argparse
+
+parser = argparse.ArgumentParser()
+parser.add_argument("--site", default="BARC", type=str,
+                    help="Denotes which site to use.")
+parser.add_argument("--target", default="oxygen", type=str,
+                    help="Specify which target time series to train on "+\
+                    "[oxygen, temperature, chla].")
+parser.add_argument("--bucket", default='shared-neon4cast-darts', type=str,
+                    help="Bucket name to connect to.")
+parser.add_argument("--endpoint", default='https://minio.carlboettiger.info', type=str,
+                    help="S3 Endpoint.")
+parser.add_argument("--accesskey", default='credentials.json', type=str,
+                    help="JSON file with access key for bucket (if required).")
+args = parser.parse_args()
+
+start = time.time()
+
+if __name__ == "__main__":
+    try:
+        s3_client = establish_s3_connection(
+            endpoint=args.endpoint,
+            json_file=args.accesskey,
+        )
+        s3_dict = {'client': s3_client, 'bucket': args.bucket}
+    except:
+        s3_client, s3_dict = None, None
+
+    preprocessor = TimeSeriesPreprocessor(
+        input_csv_name = "aquatics-targets.csv.gz",
+        load_dir_name = "preprocessed_validate/",
+        s3_dict=s3_dict,
+    )
+    preprocessor.load(args.site)
+
+    output_name = f"{args.site}/{args.target}/AutoTheta/model_0/"
+
+    model = AutoThetaForecaster(
+        validate_preprocessor=preprocessor,
+        target_variable=args.target,
+        site_id=args.site,
+        output_name=output_name,
+        s3_dict=s3_dict,
+    )
+    model.make_forecasts()
+
+    print(f"Runtime for AutoTheta on {args.target} at {args.site}: {(time.time() - start)/60:.2f} minutes")

### 2024-04-19 stats_utils.py

@@ -1,0 +1,101 @@
+import pandas as pd
+import numpy as np
+from darts.models import StatsForecastAutoTheta
+from typing import Optional
+import os
+from datetime import datetime, timedelta
+from s3_utils import (
+    upload_df_to_s3,
+)
+
+class AutoThetaForecaster():
+    def __init__(self,
+                 validate_preprocessor: Optional = None,
+                 target_variable: Optional[str] = None,
+                 season_length: Optional[int] = 365,
+                 input_chunk_length: Optional[int] = 365*5,
+                 output_name: Optional[str] = "default",
+                 forecast_horizon: Optional[int] = 30,
+                 site_id: Optional[str] = None,
+                 num_samples: Optional[int] = 500,
+                 targets_csv: Optional[str] = "aquatics-targets.csv.gz",
+                 s3_dict: Optional[dict] = {'client': None, 'bucket': None},
+                 ):
+        self.validate_preprocessor = validate_preprocessor
+        self.s3_dict = s3_dict
+        self.season_length = season_length
+        self.input_chunk_length = input_chunk_length
+        self.target_variable = target_variable
+        self.output_name = output_name
+        self.forecast_horizon = forecast_horizon
+        self.site_id = site_id
+        self.num_samples = num_samples
+        self.targets = pd.read_csv(targets_csv)
+        most_recent_date_str = np.sort(self.targets['datetime'].unique())[-1]
+        most_recent_date = datetime.strptime(most_recent_date_str, '%Y-%m-%d')
+        one_year_before = most_recent_date - timedelta(days=365)
+        self.split_date = one_year_before.strftime('%Y-%m-%d')
+
+        self.inputs = self._preprocess_data(validate_preprocessor)
+    
+        if not s3_dict['client']:
+            # Handling csv names and directories for the final forecast
+            if not os.path.exists(f"forecasts/{args.site}/{args.target}/"):
+                os.makedirs(f"forecasts/{args.site}/{args.target}/")
+        
+    def _preprocess_data(self, data_preprocessor):
+        """
+        Performs gap filling and processing of data into format that
+        Darts models will accept; train_set flag is to 
+        """
+        stitched_series_dict = data_preprocessor.sites_dict[self.site_id]
+
+        # If there was failure when filtering then we can't do preprocessing
+        if self.target_variable in \
+                data_preprocessor.site_missing_variables:
+            return "Cannot fit this target time series as no GP fit was performed."
+        inputs = stitched_series_dict[self.target_variable]
+            
+        return inputs.median()
+
[…]

### 2024-04-19 train_tune.py

@@ -6,6 +6,7 @@ from utils import (
     TimeSeriesPreprocessor,
     handle_nn_architecture,
     establish_s3_connection,
+    read_and_pivot_csv,
 )
 import argparse
 import pandas as pd
@@ -41,8 +42,6 @@ parser.add_argument("--test", default=True, action="store_false",
                     "from being saved.")
 parser.add_argument("--device", default=0, type=int,
                     help="Specify which GPU device to use [0,1].")
-parser.add_argument("--prefix", default=None, type=str,
-                    help="Prefix to use with the output csv of the forecast.")
 parser.add_argument("--bucket", default='shared-neon4cast-darts', type=str,
                     help="Bucket name to connect to.")
 parser.add_argument("--endpoint", default='https://minio.carlboettiger.info', type=str,
@@ -70,10 +69,9 @@ if __name__ == "__main__":
         
     # Selecting the device
     os.environ["CUDA_VISIBLE_DEVICES"] = f"{args.device}"
-    csv_name = "aquatics-targets.csv.gz"
 
     # Accessing the validation split date from targets csv
-    targets = pd.read_csv(csv_name)
+    targets = read_and_pivot_csv()
     most_recent_date_str = np.sort(targets['datetime'].unique())[-1]
     most_recent_date = datetime.strptime(most_recent_date_str, '%Y-%m-%d')
     one_year_before = most_recent_date - timedelta(days=365)
@@ -99,7 +97,7 @@ if __name__ == "__main__":
     preprocessors = []
     for suffix in ['train', 'validate']:
         preprocessor = TimeSeriesPreprocessor(
-            input_csv_name = csv_name,
+            input_csv_name = 'aquatics-targets.csv.gz',
             load_dir_name = f"preprocessed_{suffix}/",
             s3_dict=s3_dict,
         )
@@ -108,8 +106,6 @@ if __name__ == "__main__":
     
     
     output_name = f"{args.site}/{args.target}/{args.model}"
-    if args.prefix:
-        output_name += f"{args.prefix}"
     
     # Instantiating the model
     extras = {"epochs": args.epochs,

### 2024-04-19 utils.py

@@ -35,6 +35,20 @@ from s3_utils import(
     ls_bucket,
 )
 
+def read_and_pivot_csv(csv_name='aquatics-targets.csv.gz'):
+    '''
+    Read the targets csv and pivot wider
+    '''
+    targets_df = pd.read_csv(csv_name)
+    targets_df = targets_df.pivot_table(
+        index=['datetime', 'site_id'], 
+        columns='variable', 
+        values='observation',
+    )
+    targets_df.reset_index(inplace=True)
+
+    return targets_df
+
 def establish_s3_connection(
     endpoint='https://minio.carlboettiger.info', 
     json_file='credentials.json'):

### 2024-04-19 jobs/preprocess_timeseries.py
Need to re-run as I was using wrong targets csv

@@ -1,6 +1,7 @@
 from utils import (
     TimeSeriesPreprocessor,
     establish_s3_connection,
+    read_and_pivot_csv,
 )
 import pandas as pd
 import os
@@ -26,7 +27,7 @@ if __name__=="__main__":
     # Flag which GPU to use
     os.environ["CUDA_VISIBLE_DEVICES"] = f"{args.device}"
 
-    targets = pd.read_csv("aquatics-targets.csv.gz")
+    targets = read_and_pivot_csv()
 
     # Make s3 bucket connection
     try:

### 2024-04-19 plotting_utils.py

@@ -339,7 +339,12 @@ def modify_score_dict(csv,
     
     return score_dict
 
-def score_improvement_bysite(model, id, targets_df, target_variable, suffix="", s3_dict={'client': None, 'bucket': None}):
+def score_improvement_bysite(model, 
+                             id_, 
+                             targets_df, 
+                             target_variable, 
+                             suffix="", 
+                             s3_dict={'client': None, 'bucket': None}):
     '''
     This function collects the forecast scores for the specifed model and target variable.
     Then it returns a dataframe with columns for the difference in CRPS and RMSE
@@ -353,14 +358,14 @@ def score_improvement_bysite(model, id, targets_df, target_variable, suffix="", 
         if s3_dict['client']:
             try:
                 csv_list = ls_bucket(
-                    f'forecasts/{site_id}/{target_variable}/{model}/model_{id}/', 
+                    f'forecasts/{site_id}/{target_variable}/{model}/model_{id_}/', 
                     s3_dict, 
                     plotting=True,
                 )
             except:
                 csv_list = []
         else:
-            glob_prefix = f'forecasts/{site_id}/{target_variable}/{model}/model_{id}/*.csv'
+            glob_prefix = f'forecasts/{site_id}/{target_variable}/{model}/model_{id_}/*.csv'
             csv_list = sorted(glob.glob(glob_prefix))
         for csv in csv_list:
             site_dict = modify_score_dict(
@@ -509,12 +514,19 @@ def score_improvement_bysite(model, id, targets_df, target_variable, suffix="", 
     )
     merged_df['model'] = model
     intra_merged['model'] = model
-    merged_df['model_id'] = id
-    intra_merged['model_id'] = id
+    merged_df['model_id'] = id_
+    intra_merged['model_id'] = id_
 
     return merged_df, intra_merged
     
-def plot_forecast(date, targets_df, site_id, target_variable, model, id_list, s3_dict={'client': None, 'bucket': None}, png_name=None):
+def plot_forecast(date, 
+                  targets_df, 
+                  site_id, 
+                  target_variable, 
+                  model, 
+                  id_list, 
+                  s3_dict={'client': None, 'bucket': None}, 
+                  png_name=None):
     '''
     Returns a plot of the forecast specified by the date and model directory
     in addition to the observed values, the climatology forecast and the naive persistence
[…]

### 2024-04-19 save_plots.py

@@ -14,7 +14,7 @@ from plotting_utils import (
 import pandas as pd
 from utils import (
     establish_s3_connection,
-    NaiveEnsembleForecaster
+    NaiveEnsembleForecaster,
 )
 from s3_utils import (
     download_df_from_s3,
@@ -38,7 +38,6 @@ args = parser.parse_args()
 # Ignore all warnings; admittedly not the best practice here :/
 warnings.filterwarnings("ignore")
 
-targets = pd.read_csv("targets.csv.gz")
 metadata = generate_metadata_df()
 
 # Make s3 bucket connection
@@ -57,7 +56,7 @@ best_models_listform = {}
 target_variables = ['oxygen', 'temperature', 'chla']
 model_names = [
     "BlockRNN", "Transformer", "NBEATS", 
-    "TCN", "RNN", "TFT", 
+    "TCN", "RNN", "TFT", "AutoTheta",
     "NLinear", "DLinear", "NaiveEnsemble",
 ]
 

### 2024-04-19 save_score_dfs.py

@@ -4,7 +4,8 @@ from plotting_utils import (
 import pandas as pd
 from utils import (
     establish_s3_connection,
-    NaiveEnsembleForecaster
+    NaiveEnsembleForecaster,
+    read_and_pivot_csv
 )
 from s3_utils import (
     upload_df_to_s3,
@@ -38,7 +39,7 @@ def return_best_id(df):
 # Ignore all warnings; admittedly not the best practice here :/
 warnings.filterwarnings("ignore")
 # Note might want to change this to updated csv
-targets = pd.read_csv("targets.csv.gz")
+targets = read_and_pivot_csv()
 
 try:
     # Make s3 bucket connection
@@ -54,20 +55,20 @@ except:
 model_names = [
     "BlockRNN", "Transformer", 
     "NBEATS", "TCN", "RNN", "TFT", 
-    "NLinear", "DLinear",
+    "NLinear", "DLinear", "AutoTheta"
 ]
 target_variables = ["oxygen", "temperature", "chla"]
-ids = [0, 1, 2, 3, 4]
 s3_dict = {'client': s3_client, 'bucket': args.bucket}
 scores_dict = {}
 for model in model_names:
     scores_dict[model] = {}
-    for id_ in ids:
+    for id_ in range(5):
         # Did not train more than 2 [.]Linear models, so need to
         # treat these separately
-        if model == "NLinear" or model == "DLinear":
-            if id_ > 1:
-                continue
+        if (model == "NLinear" or model == "DLinear") and id_ > 1:
+            continue
+        if model == 'AutoTheta' and id_ > 0:
+            continue
         scores_dict[model][id_] = {}
         for target_variable in target_variables:
             inter_merged, intra_merged = score_improvement_bysite(
@@ -89,10 +90,11 @@ for target_variable in target_variables:
     for pos in ['inter', 'intra']:
         concat_list = []
         for model in model_names:
-            for id_ in ids:
-                if model == "DLinear" or model == "NLinear":
-                    if id_ > 1:
-                        continue
+            for id_ in range(5):
+                if (model == "DLinear" or model == "NLinear") and id_ > 1:
+                    continue
+                if model == "AutoTheta" and id_ > 0:
+                    continue
                 concat_list.append(scores_dict[model][id_][target_variable][pos])
         global_dfs[target_variable][pos] = pd.concat(concat_list)
 
@@ -128,10 +130,12 @@ for target_variable in target_variables:
     )
 
 # Don't need to run this cell if the dataframes have been loaded
[…]

### 2024-04-19 training_scripts/block_rnn.py
More cleaning

@@ -1,85 +1,0 @@
-from darts.utils.likelihood_models import QuantileRegression
-from residual_learning.utils import (
-                BaseForecaster, 
-                ResidualForecasterDarts,
-                TimeSeriesPreprocessor,
-)
-import argparse
-import time
-import os
-import copy
-
-os.environ["CUDA_VISIBLE_DEVICES"] = "1"
-
-start = time.time()
-parser = argparse.ArgumentParser()
-parser.add_argument("--model", default="BlockRNN", type=str)
-parser.add_argument("--target", default="oxygen", type=str)
-parser.add_argument("--site", default="BARC", type=str)
-parser.add_argument("--tune", default=False, action="store_true")
-args = parser.parse_args()
-
-data_preprocessor = TimeSeriesPreprocessor(input_csv_name = "../targets.csv.gz",
-                                           load_dir_name = "../preprocessed_timeseries/")
-data_preprocessor.load()
-
-output_csv = copy.copy(args.model)
-if args.tune:
-    output_csv += "_tuned"
-
-forecaster = BaseForecaster(model=args.model,
-                    target_variable_column_name=args.target,
-                    data_preprocessor=data_preprocessor,
-                    datetime_column_name="datetime",
-                    covariates_names=["air_tmp", "chla", "temperature"],
-                    output_csv_name=f"forecasts/{output_csv}.csv",
-                    validation_split_date="2023-02-26",
-                    model_hyperparameters={'input_chunk_length': 180},
-                    model_likelihood={"likelihood": QuantileRegression([0.05, 0.1, 0.5, 0.9, 0.95])},
-                    forecast_horizon=30,
-                    site_id=args.site)
-
-if args.tune:
-    if args.model == "BlockRNN":
-        forecaster.tune({
-            "input_chunk_length": [60, 180, 360, 540],
-            "hidden_dim": [16, 32, 64, 126],
-            "model": ["RNN", "GRU", "LSTM"],
-            "n_rnn_layers": [2, 3, 4, 5],
-        })
-    elif self.model == "TCN":
-        forecaster.tune({
-            "" : [],
-        })
-    elif self.model == "RNN":
-        forecaster.tune({
-            "" : [],
-        })
-    elif self.model == "Transformer":
-        forecaster.tune({
-            "" : [],
-        })
-    elif self.model == "NLinear":
-        forecaster.tune({
-            "" : [],
-        })
-    elif self.model == "DLinear":
-        forecaster.tune({
[…]

### 2024-04-20 aquatics_csv.R
Get targets is back

@@ -1,26 +1,0 @@
-library(tidyverse)
-library(arrow)
-library(fable)
-
-neon <- s3_bucket("neon4cast-targets/neon",
-                  endpoint_override = "data.ecoforecast.org",
-                  anonymous = TRUE)
-
-
-# we select the RH table by using its full name, this also contains air temp
-# for aquatics sites
-remote_taat <- open_dataset(neon$path("RH_30min-basic-DP1.00098.001"))
-
-
-neon_temp <- remote_taat |>
-  mutate(time = as.Date(startDateTime)) |>
-  group_by(siteID, time) |>
-  summarise(air_tmp = mean(tempRHMean, na.rm = TRUE)) |>
-  rename(site_id = siteID) |>
-  rename(datetime = time) |>
-  collect()
-
-target <- read_csv("https://data.ecoforecast.org/neon4cast-targets/aquatics/aquatics-targets.csv.gz") |>
-  pivot_wider(names_from = variable, values_from = observation) |>
-  left_join(neon_temp, by = c("site_id", "datetime")) |>
-  write_csv("targets.csv.gz")

### 2024-04-20 get_targets_csv.R

@@ -1,0 +1,28 @@
+library(dplyr)
+library(tidyverse)
+library(neon4cast)
+library(lubridate)
+
+remotes::install_github("eco4cast/neon4cast")
+
+weather_stage3 <- neon4cast::noaa_stage3()
+
+targets <- read_csv('https://data.ecoforecast.org/neon4cast-targets/aquatics/aquatics-targets.csv.gz') |>
+  pivot_wider(names_from = variable, values_from = observation)
+
+site_ids <- targets |> distinct(site_id)
+
+air_tmp <- weather_stage3 |> 
+  filter(variable == "air_temperature", site_id %in% site_ids$site_id) |>
+  mutate(datetime = as_date(datetime)) |>
+  group_by(site_id, variable, datetime) |>
+  summarize(observation = mean(prediction, na.rm = TRUE)) |>
+  mutate(observation = observation - 273.15) |>
+  collect() |>
+  pivot_wider(names_from = variable, values_from = observation)
+
+merged_data <- air_tmp |>
+  left_join(targets, by = c("site_id", "datetime")) |>
+  rename(air_tmp = air_temperature)
+
+write_csv(merged_data, path = "aquatics-targets.csv.gz")

### 2024-04-21 jobs/preprocess_timeseries.py
Identified bug but not sure why it happened, hmph

@@ -1,7 +1,6 @@
 from utils import (
     TimeSeriesPreprocessor,
     establish_s3_connection,
-    read_and_pivot_csv,
 )
 import pandas as pd
 import os
@@ -27,7 +26,7 @@ if __name__=="__main__":
     # Flag which GPU to use
     os.environ["CUDA_VISIBLE_DEVICES"] = f"{args.device}"
 
-    targets = read_and_pivot_csv()
+    targets = pd.read_csv('aquatics-targets.csv.gz')
 
     # Make s3 bucket connection
     try:

### 2024-04-21 save_score_dfs.py

@@ -4,8 +4,7 @@ from plotting_utils import (
 import pandas as pd
 from utils import (
     establish_s3_connection,
-    NaiveEnsembleForecaster,
-    read_and_pivot_csv
+    NaiveEnsembleForecaster
 )
 from s3_utils import (
     upload_df_to_s3,
@@ -39,7 +38,7 @@ def return_best_id(df):
 # Ignore all warnings; admittedly not the best practice here :/
 warnings.filterwarnings("ignore")
 # Note might want to change this to updated csv
-targets = read_and_pivot_csv()
+targets = pd.read_csv("aquatics-targets.csv.gz")
 
 try:
     # Make s3 bucket connection

### 2024-04-21 train_tune.py

@@ -6,7 +6,6 @@ from utils import (
     TimeSeriesPreprocessor,
     handle_nn_architecture,
     establish_s3_connection,
-    read_and_pivot_csv,
 )
 import argparse
 import pandas as pd
@@ -71,7 +70,7 @@ if __name__ == "__main__":
     os.environ["CUDA_VISIBLE_DEVICES"] = f"{args.device}"
 
     # Accessing the validation split date from targets csv
-    targets = read_and_pivot_csv()
+    targets = pd.read_csv("aquatics-targets.csv.gz")
     most_recent_date_str = np.sort(targets['datetime'].unique())[-1]
     most_recent_date = datetime.strptime(most_recent_date_str, '%Y-%m-%d')
     one_year_before = most_recent_date - timedelta(days=365)

### 2024-04-21 utils.py

@@ -35,20 +35,6 @@ from s3_utils import(
     ls_bucket,
 )
 
-def read_and_pivot_csv(csv_name='aquatics-targets.csv.gz'):
-    '''
-    Read the targets csv and pivot wider
-    '''
-    targets_df = pd.read_csv(csv_name)
-    targets_df = targets_df.pivot_table(
-        index=['datetime', 'site_id'], 
-        columns='variable', 
-        values='observation',
-    )
-    targets_df.reset_index(inplace=True)
-
-    return targets_df
-
 def establish_s3_connection(
     endpoint='https://minio.carlboettiger.info', 
     json_file='credentials.json'):
@@ -210,7 +196,7 @@ class HistoricalForecaster():
                  site_id: Optional[str] = None,
                  target_variable: Optional[str] = "oxygen",
                  output_csv_name: Optional[str] = "historical_forecaster_output.csv",
-                 validation_split_date: Optional[str] = "2023-03-09", #YYYY-MM-DD
+                 validation_split_date: Optional[str] = None, #YYYY-MM-DD
                  forecast_horizon: Optional[int] = 30,
                  ):
         self.targets = targets
@@ -357,7 +343,7 @@ class TimeSeriesPreprocessor():
                  s3_dict: Optional[dict] = {'client': None, 'bucket': None},
                  load_dir_name: Optional[str] = "preprocessed_timeseries/",
                  datetime_column_name: Optional[str] = "datetime",
-                 validation_split_date: Optional[str] = "2023-03-09",
+                 validation_split_date: Optional[str] = None,
                  filter_kw_args: Optional[dict] = {"alpha_0": 0.001,
                                                    "n_restarts_0": 100,
                                                    "num_samples": 500,},
@@ -372,7 +358,7 @@ class TimeSeriesPreprocessor():
         month = int(validation_split_date[5:7])
         day = int(validation_split_date[8:])
         self.split_date = pd.Timestamp(year=self.year, month=month, day=day)
-        self.df = pd.read_csv(self.input_csv_name)
+        self.df = pd.read_csv(input_csv_name)
         self.df['datetime'] = pd.to_datetime(self.df.datetime)
         self.df = self.df[self.df.datetime <= self.split_date]
     
@@ -500,8 +486,10 @@ class TimeSeriesPreprocessor():
 
         self.make_doy_dict(site_df)
         variable_list = ["chla", "oxygen", "temperature", "air_tmp"]
[…]

### 2024-04-21 jobs/test_preprocess_timeseries.py
Timeseries preprocessor back working

@@ -11,6 +11,7 @@ from darts import TimeSeries
 import numpy as np
 import argparse
 import time
+from datetime import datetime, timedelta
 
 start_ = time.time()
 
@@ -26,7 +27,7 @@ args = parser.parse_args()
 if __name__=="__main__":
     start = time.time()
     os.environ["CUDA_VISIBLE_DEVICES"] = "1"
-    targets = pd.read_csv("targets.csv.gz")
+    targets = pd.read_csv("aquatics-targets.csv.gz")
 
     # Make s3 bucket connection
     try:
@@ -38,9 +39,17 @@ if __name__=="__main__":
     except:
         s3_dict = {'client': None, 'bucket': None}
 
+    # Defining the training data split date to be one year before the most
+    # date time in aquatics-targets.csv.gz. Validation split date will be
+    # the most recent date.
+    most_recent_date_str = np.sort(targets['datetime'].unique())[-1]
+    most_recent_date = datetime.strptime(most_recent_date_str, '%Y-%m-%d')
+    one_year_before = most_recent_date - timedelta(days=365)
+    one_year_before_str = one_year_before.strftime('%Y-%m-%d')
+
     # For the training set
     data_preprocessor = TimeSeriesPreprocessor(
-        validation_split_date='2022-07-19',
+        validation_split_date=one_year_before_str,
         load_dir_name='preprocessed_train/',
         s3_dict=s3_dict,
     )

### 2024-04-21 utils.py

@@ -354,10 +354,8 @@ class TimeSeriesPreprocessor():
         self.filter_kw_args = filter_kw_args
         self.sites_dict = {}
         self.s3_dict = s3_dict
-        self.year = int(validation_split_date[:4])
-        month = int(validation_split_date[5:7])
-        day = int(validation_split_date[8:])
-        self.split_date = pd.Timestamp(year=self.year, month=month, day=day)
+        self.split_date = pd.to_datetime(validation_split_date)
+        self.year = self.split_date.year
         self.df = pd.read_csv(input_csv_name)
         self.df['datetime'] = pd.to_datetime(self.df.datetime)
         self.df = self.df[self.df.datetime <= self.split_date]
@@ -468,6 +466,7 @@ class TimeSeriesPreprocessor():
         self.sites_dict_null = {}
         # Preparing a dataframe
         site_df = self.df.loc[self.df.site_id == site]
+        site_df = site_df.sort_values(by=self.datetime_column_name)
         times = pd.to_datetime(site_df[self.datetime_column_name])
         times = pd.DatetimeIndex(times)
 
@@ -486,10 +485,9 @@ class TimeSeriesPreprocessor():
 
         self.make_doy_dict(site_df)
         variable_list = ["chla", "oxygen", "temperature", "air_tmp"]
-        # I am not exactly sure why but there are duplicate time indices, so I need to remove them
-        # when creating the time series
-        self.var_tseries_dict = {var: TimeSeries.from_times_and_values(times[~times.duplicated()], 
-                                                                 site_df[[var]][~times.duplicated()], 
+        # I think this is coming from above
+        self.var_tseries_dict = {var: TimeSeries.from_times_and_values(times, 
+                                                                 site_df[[var]], 
                                                                  fill_missing_dates=True,
                                                                  freq="D") 
                                                         for var in variable_list}

### 2024-04-24 train_tune.py
Checking changes

@@ -10,6 +10,7 @@ from utils import (
 import argparse
 import pandas as pd
 import time
+import sys
 import os
 import copy
 import json
@@ -95,7 +96,7 @@ if __name__ == "__main__":
     # Loading data preprocessors for training and validation
     preprocessors = []
     for suffix in ['train', 'validate']:
-        split_date = most_recent_date_str if suffix == 'train' \
+        split_date = most_recent_date_str if suffix == 'validate' \
                         else validation_split_date
         preprocessor = TimeSeriesPreprocessor(
             input_csv_name = 'aquatics-targets.csv.gz',
@@ -105,7 +106,11 @@ if __name__ == "__main__":
         )
         preprocessor.load(args.site)
         preprocessors.append(preprocessor)
-    
+
+    if preprocessors[0].site_missing_variables != \
+         preprocessors[1].site_missing_variables:
+        print("Missing data edge case. Training not performed.")
+        sys.exit()
     
     output_name = f"{args.site}/{args.target}/{args.model}"
     
@@ -120,6 +125,7 @@ if __name__ == "__main__":
             **hyperparams_dict["model_hyperparameters"],
             **nn_arg,
         }
+        
         forecaster = BaseForecaster(
             model=args.model,
             target_variable=args.target,

### 2024-04-24 utils.py

@@ -686,15 +686,10 @@ class BaseForecaster():
             covariates = stitched_series_dict[self.covariates_names[0]]
             for cov_var in self.covariates_names[1:]:
                 covariates = covariates.concatenate(stitched_series_dict[cov_var], 
-                                                              axis=1, 
-                                                              ignore_time_axis=True)
+                                                    axis=1, 
+                                                    ignore_time_axis=True)
             covariates = covariates.median()
-
-            covs_train, _ = (
-                covariates
-                .median()
-                .split_after(self.split_date)
-            )
+            covs_train, _ = covariates.split_after(self.split_date)
         else:
             covs_train = None
             covariates = None

### 2024-04-24 predict_stats.py
Ignoring problematic sites in this script as well

@@ -37,19 +37,36 @@ if __name__ == "__main__":
     # Accessing the validation split date from targets csv
     targets = pd.read_csv("aquatics-targets.csv.gz")
     most_recent_date_str = np.sort(targets['datetime'].unique())[-1]
+    most_recent_date = datetime.strptime(most_recent_date_str, '%Y-%m-%d')
+    one_year_before = most_recent_date - timedelta(days=365)
+    validation_split_date = one_year_before.strftime('%Y-%m-%d')
     
-    preprocessor = TimeSeriesPreprocessor(
-        input_csv_name = "aquatics-targets.csv.gz",
-        load_dir_name = "preprocessed_validate/",
-        s3_dict=s3_dict,
-        validation_split_date=most_recent_date_str,
-    )
-    preprocessor.load(args.site)
+    # Loading data preprocessors for training and validation
+    # Note that having different preprocessors is not required
+    # for the Theta model, but I am still loading both as to
+    # see what site to train on.
+    preprocessors = []
+    for suffix in ['train', 'validate']:
+        split_date = most_recent_date_str if suffix == 'validate' \
+                        else validation_split_date
+        preprocessor = TimeSeriesPreprocessor(
+            input_csv_name = 'aquatics-targets.csv.gz',
+            load_dir_name = f"preprocessed_{suffix}/",
+            s3_dict=s3_dict,
+            validation_split_date=split_date,
+        )
+        preprocessor.load(args.site)
+        preprocessors.append(preprocessor)
+
+    if preprocessors[0].site_missing_variables != \
+         preprocessors[1].site_missing_variables:
+        print("Missing data edge case. Training not performed.")
+        sys.exit()
 
     output_name = f"{args.site}/{args.target}/AutoTheta/model_0/"
 
     model = AutoThetaForecaster(
-        validate_preprocessor=preprocessor,
+        validate_preprocessor=preprocessors[1],
         target_variable=args.target,
         site_id=args.site,
         output_name=output_name,

### 2024-04-24 predict_stats.py
Forgot an import

@@ -7,6 +7,8 @@ import pandas as pd
 import numpy as np
 import time
 import argparse
+from datetime import datetime, timedelta
+import sys
 
 parser = argparse.ArgumentParser()
 parser.add_argument("--site", default="BARC", type=str,

### 2024-04-24 jobs/preprocess_timeseries.py
Hmm

@@ -47,25 +47,25 @@ if __name__=="__main__":
     one_year_before_str = one_year_before.strftime('%Y-%m-%d')
 
     # For the training set
-    data_preprocessor = TimeSeriesPreprocessor(
+    train_preprocessor = TimeSeriesPreprocessor(
         validation_split_date=one_year_before_str,
         load_dir_name='preprocessed_train/',
         s3_dict=s3_dict,
     )
     
-    _ = [data_preprocessor.preprocess_data(site) for site in targets.site_id.unique()]
+    _ = [train_preprocessor.preprocess_data(site) for site in targets.site_id.unique()]
     
-    data_preprocessor.save()
 
     # For the validation set
-    data_preprocessor = TimeSeriesPreprocessor(
+    validate_preprocessor = TimeSeriesPreprocessor(
         validation_split_date=most_recent_date_str,
         load_dir_name='preprocessed_validate/',
         s3_dict=s3_dict,
     )
     
-    _ = [data_preprocessor.preprocess_data(site) for site in targets.site_id.unique()]
-    
+    _ = [validate_preprocessor.preprocess_data(site) for site in targets.site_id.unique()]
+
+    train_preprocessor.save()
     data_preprocessor.save()
 
     

### 2024-04-25 plotting_utils.py
Plots are looking pretty now :sunglasses:

@@ -685,13 +685,12 @@ def plot_improvement_bysite(score_df, metadata_df, historical=True, png_name=Non
     Returns a plot of the scoring metric difference vs. the site id;
     site type is encoded by color.
     '''
-    ## Find the percentage of forecast windows during which the ML model excelled 
-    ## the historical forecaster
-    column = (
-        'skill_historical_ml_crps' if historical \
-         else 'skill_naive_ml_rmse'
-    )
-
+    plt.figure(figsize=(12, 8))
+    color_dict = {
+        'Wadeable Stream': 'tab:blue', 
+        'Lake': 'indianred', 
+        'Non-wadeable River': 'plum'
+    }
     # Combine df's to include metadata
     df = pd.merge(
         score_df, 
@@ -699,25 +698,28 @@ def plot_improvement_bysite(score_df, metadata_df, historical=True, png_name=Non
         right_on='field_site_id', 
         left_on='site_id'
     ).drop(columns=['field_site_id'])
+
+    ## Find the percentage of forecast windows during which the ML model excelled 
+    ## the historical forecaster
+    column = (
+        'skill_historical_ml_crps' if historical \
+         else 'skill_naive_ml_rmse'
+    )
+    df['transformed_column'] = np.abs(df[column] - 1)
+    column = 'transformed_column'
     
-    plt.figure(figsize=(12, 8))
-    color_dict = {
-        'Wadeable Stream': 'tab:blue', 
-        'Lake': 'indianred', 
-        'Non-wadeable River': 'plum'
-    }
 
     for site_type in ['Wadeable Stream', 'Lake', 'Non-wadeable River']:
-        sns.boxplot(
+        sns.stripplot(
             data=df.loc[df.field_site_subtype == site_type],
             x='site_id',
             y=column,
             color=color_dict[site_type],
-            showfliers=False,
+            log_scale=2,
         )
 
     plt.grid(False)
-    plt.axhline(y=0, color='black', linestyle='dashed', linewidth=1)
+    plt.axhline(y=1, color='black', linestyle='dashed', linewidth=1)
     if historical:
         plt.ylabel("CRPSS")
     else:
@@ -743,22 +745,26 @@ def plot_global_percentages(df_, historical=True, png_name=None):
     column = (
         'skill_historical_ml_crps' if historical \
          else 'skill_naive_ml_rmse'
-    ) 
+    )
[…]

### 2024-04-26 plotting_utils.py
Coming along!

@@ -863,28 +863,61 @@ def plot_site_type_percentages_bymodel(df_, metadata_df, historical=True, png_na
         dodge=True,
         palette=color_dict,
         log_scale=2,
+        legend=False,
     )
 
     plt.grid(False)
     plt.axhline(y=1, color='black', linestyle='dashed', linewidth=1)
     if historical:
-        plt.ylabel("|CRPSS - 1|", fontsize=18)
+        plt.ylabel("|CRPSS - 1|", fontsize=30)
     else:
-        plt.ylabel("|RMSE-SS - 1|", fontsize=18)
+        plt.ylabel("|RMSE-SS - 1|", fontsize=30)
     ax = plt.gca()
     ax.spines["left"].set_visible(True)
     ax.spines["bottom"].set_visible(True)
-    plt.xlabel("model", fontsize=18)
-    plt.xticks(rotation=30, fontsize=14)
-    plt.yticks(fontsize=14)
-    legend_handles = [Patch(facecolor=color, edgecolor='black') for color in color_dict.values()]
-    legend_labels = list(color_dict.keys())
-    plt.legend(legend_handles, legend_labels, title='Site Type', loc='upper right')
+    plt.xlabel("model", fontsize=30)
+    plt.xticks(rotation=30, fontsize=24)
+    plt.yticks(fontsize=24)
     plt.tight_layout()
 
     # Saving the plot if desired
     if png_name:
         save_fig(plt, png_name)
+
+    plt.clf()
+
+    # Create a figure and axis for the legend plot
+    legend_fig, legend_ax = plt.subplots(figsize=(26,2))  # Adjust size as needed
+    
+    # Extract model names and colors from color_palette
+    models = list(color_dict.keys())
+    colors = list(color_dict.values())
+    
+    # Plot lines for each model with corresponding colors
+    legend_lines = [Line2D([0], [0], marker='o', color=color, markersize=22, linestyle='') for color in colors]
+    
+    # Create legend with circles and model names, orient horizontally with 1 column
+    legend_ax.legend(
+        legend_lines, 
+        models, 
+        loc='center', 
+        fontsize=30, 
+        ncol=len(models), 
+        handlelength=1,
+        handletextpad=0.02,
+        bbox_to_anchor=(0.5, 0.5),
+        labelspacing=4
+    )
+    
+    # Customize legend appearance
+    legend_ax.axis('off')  # Hide axis
+    
+    # Show the plot
+    plt.tight_layout()
+    
+    # Saving the legend plot if desired
[…]

### 2024-04-26 manuscript/manuscript.Rmd
Made some edits to Theta Forecaster; need to re-run that now

@@
 ---
-title: "A Comparison of Machine Learning Models for Water Quality Forecasting"
+title: "A Comparison of Neural Network Models for Water Quality Forecasting"
 authors:
@@
     \centering
+    \includegraphics[width=\textwidth]{../plots/sitetype_legend.png}
+  \end{subfigure}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
     \includegraphics[width=\textwidth]{../plots/type_local_historical_oxygen.png}
   \end{subfigure}
-
-  \begin{subfigure}{\textwidth}
+  \begin{subfigure}{0.49\textwidth}
     \centering
@@
   \end{subfigure}
-
   \caption{Comparison of Historical and Naive Oxygen}
@@
     \centering
+    \includegraphics[width=\textwidth]{../plots/sitetype_legend.png}
+  \end{subfigure}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
     \includegraphics[width=\textwidth]{../plots/type_local_historical_temperature.png}
   \end{subfigure}
-
-  \begin{subfigure}{\textwidth}
+  \begin{subfigure}{0.49\textwidth}
     \centering
@@
   \end{subfigure}
-
   \caption{Comparison of Historical and Naive temperature}
@@
     \centering
+    \includegraphics[width=\textwidth]{../plots/sitetype_legend.png}
+  \end{subfigure}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
     \includegraphics[width=\textwidth]{../plots/type_local_historical_chla.png}
   \end{subfigure}
-
-  \begin{subfigure}{\textwidth}
+  \begin{subfigure}{0.49\textwidth}
     \centering
@@
   \end{subfigure}
-
   \caption{Comparison of Historical and Naive chla}
@@
   \centering
-
   \begin{subfigure}{0.3\textwidth}
@@
   \end{subfigure}
-  \hfill
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_naive_oxygen.png}
+  \end{subfigure}
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \rule{0pt}{0pt}
+  \end{subfigure}
   \begin{subfigure}{0.3\textwidth}
@@
   \end{subfigure}
-  \hfill
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_naive_temperature.png}
+  \end{subfigure}
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/intra_legend.png}
+  \end{subfigure}
   \begin{subfigure}{0.3\textwidth}
@@
   \end{subfigure}
-
-  \begin{subfigure}{0.3\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{../plots/intra_naive_oxygen.png}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}{0.3\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{../plots/intra_naive_temperature.png}
-  \end{subfigure}
-  \hfill
   \begin{subfigure}{0.3\textwidth}
@@
     \includegraphics[width=\textwidth]{../plots/intra_naive_chla.png}
+  \end{subfigure}
+  \begin{subfigure}{0.3\textwidth}
+    \centering
+    \rule{0pt}{0pt}
   \end{subfigure}
@@
     \centering
+    \includegraphics[width=\textwidth]{../plots/sitetype_legend.png}
+  \end{subfigure}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
     \includegraphics[width=\textwidth]{../plots/global_loc_historical_oxygen.png}
   \end{subfigure}
-
-  \begin{subfigure}{\textwidth}
+  \begin{subfigure}{0.49\textwidth}
     \centering
@@
   \end{subfigure}
-
   \caption{Global performance on oxygen}

### 2024-04-26 plotting_utils.py

@@ -578,16 +578,16 @@ def plot_forecast(date,
     historical_model.forecast_ts.plot(label="Historical", color=colors[1])
     validation_series.plot(label="Truth", color=colors[2])
 
-    # And the naive forecaster
-    naive_model = NaivePersistenceForecaster(
-        targets=targets_df,
-        site_id=site_id,
-        target_variable=target_variable,
-        validation_split_date=str(model_forecast.time_index[0])[:10],
-        forecast_horizon=len(model_forecast),
-    )
-    naive_model.make_forecasts()
-    naive_model.forecast_ts.plot(label='Naive Persistence', color=colors[3])
+    ## And the naive forecaster
+    #naive_model = NaivePersistenceForecaster(
+    #    targets=targets_df,
+    #    site_id=site_id,
+    #    target_variable=target_variable,
+    #    validation_split_date=str(model_forecast.time_index[0])[:10],
+    #    forecast_horizon=len(model_forecast),
+    #)
+    #naive_model.make_forecasts()
+    #naive_model.forecast_ts.plot(label='Naive Persistence', color=colors[3])
     
     x = plt.xlabel("date")
     y = plt.ylabel(target_variable)

### 2024-04-26 stats_utils.py

@@ -12,8 +12,9 @@ class AutoThetaForecaster():
     def __init__(self,
                  validate_preprocessor: Optional = None,
                  target_variable: Optional[str] = None,
-                 season_length: Optional[int] = 365,
-                 input_chunk_length: Optional[int] = 365*5,
+                 season_length: Optional[int] = 1,
+                 decomposition_type: Optional[str] = 'additive',
+                 input_chunk_length: Optional[int] = 31,
                  output_name: Optional[str] = "default",
                  forecast_horizon: Optional[int] = 30,
                  site_id: Optional[str] = None,
@@ -24,6 +25,7 @@ class AutoThetaForecaster():
         self.validate_preprocessor = validate_preprocessor
         self.s3_dict = s3_dict
         self.season_length = season_length
+        self.decomposition_type = decomposition_type
         self.input_chunk_length = input_chunk_length
         self.target_variable = target_variable
         self.output_name = output_name
@@ -79,7 +81,10 @@ class AutoThetaForecaster():
         This function fits a Darts model to the training_set
         """
 
-        self.model = StatsForecastAutoTheta(season_length=self.season_length)
+        self.model = StatsForecastAutoTheta(
+            season_length=self.season_length,
+            decomposition_type=self.decomposition_type,
+        )
         predict_series = self.get_predict_set(self.input_chunk_length)
 
         predictions = []

### 2024-04-30 plotting_utils.py
Converging on final plots I think

@@ -23,6 +23,7 @@ from darts.metrics import rmse
 import matplotlib as mpl
 from sklearn.cluster import KMeans
 from datetime import datetime
+import random
 
 pd.options.mode.chained_assignment = None
 
@@ -532,9 +533,17 @@ def plot_forecast(date,
     in addition to the observed values, the climatology forecast and the naive persistence
     forecast.
     '''
+    #plt.figure(figsize=(12, 8))
+    
     cmap = mpl.colormaps["tab10"]
     colors = cmap.colors
-    dfs = []
+    if model == 'AutoTheta':
+        color = colors[0]
+    elif model == 'NaiveEnsemble':
+        color = colors[9]
+    elif model == 'TFT':
+        color = colors[4]
+    
     for i, id_ in enumerate(id_list):
         # Loading the forecast csv and creating a time series
         if s3_dict['client']:
@@ -552,7 +561,7 @@ def plot_forecast(date,
         model_forecast = TimeSeries.from_times_and_values(times, 
                                                           values, 
                                                           fill_missing_dates=True, freq="D")
-        model_forecast.plot(label=f"{model}", color=colors[0])
+        model_forecast.plot(color=color)
 
     # Getting the validation series directly from the targets csv
     date = model_forecast.time_index[0]
@@ -575,8 +584,8 @@ def plot_forecast(date,
         forecast_horizon=len(model_forecast),
     )
     historical_model.make_forecasts()
-    historical_model.forecast_ts.plot(label="Historical", color=colors[1])
-    validation_series.plot(label="Truth", color=colors[2])
+    historical_model.forecast_ts.plot(color=colors[1])
+    validation_series.plot(color=colors[2])
 
     ## And the naive forecaster
     #naive_model = NaivePersistenceForecaster(
@@ -591,20 +600,22 @@ def plot_forecast(date,
     
     x = plt.xlabel("date")
     y = plt.ylabel(target_variable)
+    plt.legend().remove()
     # Creating a legend and then removing duplicates
     ax = plt.gca()
-    handles, labels = plt.gca().get_legend_handles_labels()
-    unique_labels = []
-    unique_handles = []
-    for i, label in enumerate(labels):
-        if label not in unique_labels:
-            unique_labels.append(label)
-            unique_handles.append(handles[i])
-    
[…]

### 2024-04-30 save_plots.py

@@ -10,6 +10,7 @@ from plotting_utils import (
     plot_window_and_sitetype_performance,
     generate_metadata_df,
     plot_crps_over_time_agg,
+    make_forecast_legend,
 )
 import pandas as pd
 from utils import (
@@ -80,6 +81,8 @@ for target_variable in target_variables:
                 [model, int(df[df['model'] == model]['model_id'].unique())] for model in model_names
             ]
 
+# Generate legend for individual forecasts
+make_forecast_legend()
 
 for target_variable in target_variables:
     # These plots incorporate forecasts from all models

### 2024-04-30 plotting_utils.py
Creating a new branch here

@@ -483,23 +483,24 @@ def score_improvement_bysite(model,
         1 - (rmse_merged['value_historical'] / rmse_merged['value_naive'])
     )
 
+    import pdb; pdb.set_trace()
     # Delete unnecessary columns
-    rmse_merged = rmse_merged.drop(
-        rmse_merged.filter(like='model').columns, 
-        axis=1
-    )
-    rmse_merged = rmse_merged.drop(
-        rmse_merged.filter(like='value').columns, 
-        axis=1
-    )
-    crps_merged = crps_merged.drop(
-        crps_merged.filter(like='model').columns, 
-        axis=1
-    )
-    crps_merged = crps_merged.drop(
-        crps_merged.filter(like='value').columns, 
-        axis=1
-    )
+    #rmse_merged = rmse_merged.drop(
+    #    rmse_merged.filter(like='model').columns, 
+    #    axis=1
+    #)
+    #rmse_merged = rmse_merged.drop(
+    #    rmse_merged.filter(like='value').columns, 
+    #    axis=1
+    #)
+    #crps_merged = crps_merged.drop(
+    #    crps_merged.filter(like='model').columns, 
+    #    axis=1
+    #)
+    #crps_merged = crps_merged.drop(
+    #    crps_merged.filter(like='value').columns, 
+    #    axis=1
+    #)
 
     # Joining the two df's along site id and date then adding a combined improvement column
     # for comparison against the climatology model
@@ -517,6 +518,7 @@ def score_improvement_bysite(model,
     intra_merged['model'] = model
     merged_df['model_id'] = id_
     intra_merged['model_id'] = id_
+
 
     return merged_df, intra_merged
     

### 2024-04-30 s3_utils.py

@@ -32,7 +32,7 @@ def read_credentials_from_json(file_path):
 def upload_df_to_s3(object_name, dataframe, s3_dict={'client': None, 'bucket': None}):
     # Convert DataFrame to CSV format in memory
     csv_buffer = StringIO()
-    dataframe.to_csv(csv_buffer)
+    dataframe.to_csv(csv_buffer, index=False)
     
     s3_dict['client'].put_object(
         Bucket=s3_dict['bucket'], 

### 2024-04-30 save_plots.py

@@ -83,7 +83,7 @@ for target_variable in target_variables:
 
 # Generate legend for individual forecasts
 make_forecast_legend()
-
+import pdb; pdb.set_trace()
 for target_variable in target_variables:
     # These plots incorporate forecasts from all models
     plot_crps_over_time_agg(

### 2024-05-02 plotting_utils.py
Last changes before review break

@@ -483,20 +483,19 @@ def score_improvement_bysite(model,
         1 - (rmse_merged['value_historical'] / rmse_merged['value_naive'])
     )
 
-    import pdb; pdb.set_trace()
     # Delete unnecessary columns
-    #rmse_merged = rmse_merged.drop(
-    #    rmse_merged.filter(like='model').columns, 
-    #    axis=1
-    #)
+    rmse_merged = rmse_merged.drop(
+        rmse_merged.filter(like='model').columns, 
+        axis=1
+    )
     #rmse_merged = rmse_merged.drop(
     #    rmse_merged.filter(like='value').columns, 
     #    axis=1
     #)
-    #crps_merged = crps_merged.drop(
-    #    crps_merged.filter(like='model').columns, 
-    #    axis=1
-    #)
+    crps_merged = crps_merged.drop(
+        crps_merged.filter(like='model').columns, 
+        axis=1
+    )
     #crps_merged = crps_merged.drop(
     #    crps_merged.filter(like='value').columns, 
     #    axis=1
@@ -508,7 +507,8 @@ def score_improvement_bysite(model,
         crps_merged, 
         rmse_merged, 
         on=['site_id', 'date'], 
-        how='inner'
+        how='inner',
+        suffixes=('_crps', '_rmse')
     )
     merged_df = merged_df.drop(
         merged_df.filter(like='metric').columns, 
@@ -518,7 +518,6 @@ def score_improvement_bysite(model,
     intra_merged['model'] = model
     merged_df['model_id'] = id_
     intra_merged['model_id'] = id_
-
 
     return merged_df, intra_merged
     
@@ -543,7 +542,7 @@ def plot_forecast(date,
         color = colors[0]
     elif model == 'NaiveEnsemble':
         color = colors[9]
-    elif model == 'TFT':
+    else:
         color = colors[4]
     
     for i, id_ in enumerate(id_list):

### 2024-05-02 save_plots.py

@@ -83,7 +83,6 @@ for target_variable in target_variables:
 
 # Generate legend for individual forecasts
 make_forecast_legend()
-import pdb; pdb.set_trace()
 for target_variable in target_variables:
     # These plots incorporate forecasts from all models
     plot_crps_over_time_agg(

### 2024-05-02 manuscript/manuscript.Rmd
Converging on final plots I think

@@
 
+\begin{table}[!htbp]
+    \centering
+    \caption{Mean CRPS and RMSE}
+    \begin{subtable}{\textwidth}
+        \centering
+        \caption{Dissolved Oxygen}
+        \begin{tabular}{lcc}
+            \toprule
+            Model & Mean CRPS & Mean RMSE \\
+            \midrule
+            Climatology & 0.62 & 0.92 \\
+            Naive Persistence & & 1.55 \\
+            AutoTheta & 0.67 & 1.02 \\
+            BlockRNN & 0.52 & 0.80 \\
+            DLinear & 0.52 & 0.81  \\
+            NBEATS & 0.51 & 0.80 \\
+            NLinear & 0.52 & 0.80 \\
+            \textbf{NaiveEnsemble} & \textbf{0.47} & \textbf{0.74} \\
+            RNN & 0.60 & 0.92 \\
+            TCN & 0.62 & 0.98 \\
+            TFT & 0.52 & 0.80 \\
+            Transformer & 0.51 & 0.80 \\
+            \bottomrule
+        \end{tabular}
+    \end{subtable}
+    \begin{subtable}{\textwidth}
+        \centering
+        \caption{Water Temperature}
+        \begin{tabular}{lcc}
+            \toprule
+            Model & Mean CRPS & Mean RMSE \\
+            \midrule
+            Climatology & 1.46 & 2.19 \\
+            Naive Persistence & & 6.05 \\
+            AutoTheta & 1.55 & 2.35 \\
+            BlockRNN & 1.45 & 2.26 \\
+            DLinear & 1.35 & 2.08 \\
+            NBEATS & 1.36 & 2.10 \\
+            NLinear & 1.33 & 2.06 \\
+            \textbf{NaiveEnsemble} & \textbf{1.18} & \textbf{1.80} \\
+            RNN & 1.32 & 2.03 \\
+            TCN & 1.89 & 3.02 \\
+            TFT & 1.19 & 1.86 \\
+            Transformer & 1.32 & 2.05 \\
+            \bottomrule
+        \end{tabular}
+    \end{subtable}
+    \begin{subtable}{\textwidth}
+        \centering
+        \caption{Chlorophyll-a}
+        \begin{tabular}{lcc}
+            \toprule
+            Model & Mean CRPS & Mean RMSE \\
+            \midrule
+            Climatology & 4.83 & 7.50 \\
+            \textbf{Naive Persistence} & & \textbf{4.86} \\
+            AutoTheta &  4.20 & 6.31 \\
+            \textbf{BlockRNN} & \textbf{3.14} & 5.16 \\
+            DLinear & 3.47 & 5.48 \\
+            NBEATS & 3.78 & 5.95 \\
+            NLinear & 3.67 & 5.63 \\
+            NaiveEnsemble & 3.44 & 5.50 \\
+            RNN & 4.15 & 6.40 \\
+            TCN & 3.72 & 5.77 \\
+            TFT & 3.67 & 5.87 \\
+            Transformer & 4.02 & 6.35 \\
+            \bottomrule
+        \end{tabular}
+    \end{subtable}
+    \label{tab:mean_crps_rmse}
+\end{table}
+
 \begin{figure}[!htbp]
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/sitetype_legend.png}
+    \includegraphics[width=\textwidth]{../plots/sitetype_legend_0.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/sitetype_legend.png}
+    \includegraphics[width=\textwidth]{../plots/sitetype_legend_1.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/sitetype_legend.png}
+    \includegraphics[width=\textwidth]{../plots/forecast_legend.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/global_loc_historical_oxygen.png}
+    \includegraphics[width=\textwidth]{../plots/oxygen_theta.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/date_type_global_historical_oxygen.png}
+    \includegraphics[width=\textwidth]{../plots/chla_theta.png}
   \end{subfigure}
-  \caption{Global performance on oxygen}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/oxygen_tft.png}
+  \end{subfigure}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/chla_tft.png}
+  \end{subfigure}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/oxygen_ensemble.png}
+  \end{subfigure}
+  \begin{subfigure}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{../plots/chla_ensemble.png}
+  \end{subfigure}
+  \caption{Individual Forecast Examples}
 \end{figure}

### 2024-05-23 manuscript/manuscript.Rmd
Coming along

@@
 ---
+\setlength{\parindent}{15pt}
 
@@
 
-The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is important to natural resource management.
-Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to prepare for scenarios with far-reaching negative impacts.
-Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}.
-A shortcoming of the limnological studies that support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of water systems on a broad geographic scale.
-This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.
+The ability to accurately forecast water quality variables has become increasingly important in the era of global change.
+Freshwater ecosystems have been disproportionately impacted by anthropogenic activities, a trend that is expected to continue in the future \cite{albert2021}.
+Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{}.
+Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{Hanson, zwart}.
+A shortcoming of past limnological forecasting studies that support the use of machine learning is that they tend to focus on a selection of sites that are not representative of freshwater systems across a broad geographic scale.
+This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by evaluating forecasts at 34 different sites across North America at different times of the year.
 
@@
 A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}.
-For instance, in a model comparison that examined the forecasts for phenology time series, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
-Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the climatology model.
+For instance, in a model comparison that examined the forecasts for phenological time series, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
+Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the historical null model.
 
-There are many neural network architectures that have not been explored for ecological time series forecasting.
-Past applications of machine learning to limnological time series forecasting have typically used the long-short-term-memory (LSTM) neural network which was first introduced in 1997 \cite{lstm}.
-Over the nearly 30 years since LSTM's were introduced, there has been much development in the subfield of machine learning for time series forecasting, but recent studies examining machine learning for ecological time series forecasting have not explored more contemporary neural network architectures.
-This paper will compare 8 machine learning algorithms including LSTM's to ML methods that have been recently developed for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.
+There are many promising neural network architectures that have not been evaluated for water quality forecasting.
+Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network which was published in 1997 \cite{lstm}.
+Over the nearly 30 years since LSTM's were introduced, researchers have identified that LSTM's have notable shortcomings, and there have been continued efforts towards the development of better performing models.
+Many recent studies that use neural networks for water quality forecasting have not explored more contemporary neural network architectures.
+We will address this gap in the literature by comparing 8 neural network models including LSTM's as well as more recently published neural network models.
 
@@
 
-We employed ML models to generate water temperature, dissolved oxygen and chlorophyll-A concentration forecasts for 34 freshwater sites across North America.
+We evaluated the forecasting performance of 11 different models on water temperature, dissolved oxygen and chlorophyll-A time series, which were collected by in-situ sensors at 34 freshwater sites across North America.
 These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON.
-8 different ML models were trained for every site and target variable available in NEON Ecological Forecasting Challenge Aquatics theme.
+There is variability across sites in which target variables were observed.
+Additionally, maintenance issues led to gaps in the data which also varied with the site.
+Since the training data spanned less than 3 years, we excluded time series that had gaps longer than a year.
+Provided the lack of time series and large gaps at certain sites, we evaluated the forecast performance for water temperature, dissolved oxygen and chlorophyll-A at 33, 32 and 9 sites, respectively.
+
+The imputation of missing data proved to be critical to the performance of the ML models.
+After experimenting with data filling methods that resulted in poor model performance, we developed an imputation method inspired by the climatology model.
+If the gap size was less than 5 days, then the gap was filled using a Gaussian Process Filter.
+For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, for the case when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.
+
+We compared the performance of the neural network models to the climatology and naive persistence model.
+The climatology model generates forecasts by finding the daily mean and standard deviation and draws samples from a Gaussian Distribution with these parameters.
+The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast horizon.
+For each model, we computed the Continuous Ranked Probability Score (CRPS) and the Root Mean Square Error (RMSE).
+To gauge how models performed in relation to the climatology model, we computed the Continuous Ranked Probability Skill Score (CRPSS) which we defined as
+\begin{equation}
+CRPSS_{model} = 1 - \frac{CRPS_{model}}{CRPS_{clim.}}.
+\end{equation}
+
+And using the naive persistence model as a reference, we computed the RMSE Skill Score,
+
+\begin{equation}
+RMSE-SS_{model} = 1 - \frac{RMSE_{model}}{RMSE_{naive.}}.
+\end{equation}
+
+If a target model outperforms than the reference model, then the target model will have a positive skill score.
+If the target model performs worse than the reference model, the skill score will be negative.
+
+Talk about theta model
+
+For the 8 neural network models that were investigated, there were varying design choices made regarding which type of covariates could be used.
+For instance, models could accept past covariates only, future covariates only or both future and past covariates.
+
 For the models that accept past covariates (**list these models**), we used the other target variables recorded at that site as well as air temperature as covariates.
 For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate.
-The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2018 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
-
-We compared the ML models to a climatology and naive persistence null model.
-The climatology model generates forecasts by finding the daily mean and standard deviation of the target time series, and then drawing samples from a Gaussian Distribution with these parameters.
-The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast window.
-
-The imputation of missing data proved to be critical to the performance of the machine learning models.
-After experimenting with data filling methods that resulted in poor model performance, we settled on a imputation method inspired by the climatology model.
-Missing data is estimated using the daily historical mean when this statistic is available; and, for the case when there are no data collected for a day of the year, either the monthly, seasonal or global median was used with preference given to the median available from the shortest timescale.
+The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
 
@@
 
+Models tend to struggle to accurately forecast chlorophyll during bloom conditions \cite{woelmer}. Accurately predicting timing and magnitude of bloom events is not surprising
+
 # Acknowledgements

### 2024-05-23 manuscript/manuscript.Rmd
End of day updates

@@
 There is variability across sites in which target variables were observed.
-Additionally, maintenance issues led to gaps in the data which also varied with the site.
-Since the training data spanned less than 3 years, we excluded time series that had gaps longer than a year.
+Additionally, maintenance issues led to gaps in the data which also varied across locations.
 Provided the lack of time series and large gaps at certain sites, we evaluated the forecast performance for water temperature, dissolved oxygen and chlorophyll-A at 33, 32 and 9 sites, respectively.
@@
 \end{equation}
-
 And using the naive persistence model as a reference, we computed the RMSE Skill Score,
-
 \begin{equation}
@@
 \end{equation}
-
-If a target model outperforms than the reference model, then the target model will have a positive skill score.
+If a target model outperforms the reference model, then the target model will have a positive skill score.
 If the target model performs worse than the reference model, the skill score will be negative.
 
-Talk about theta model
+## Method Group 1: Statistical Models (Theta)
+
+## Method Group 2: Neural Network Models
 
@@
 When comparing the probabilistic forecasts, we express the CRPS score in error-orientation -- i.e., a lower CRPS indicates a smaller margin of error.
-
-## Machine Learning Models
 

### 2024-05-28 manuscript/manuscript.Rmd
Updating changes here
Theta paragraph up

@@
 \begin{equation}
-CRPSS_{model} = 1 - \frac{CRPS_{model}}{CRPS_{clim.}}.
+CRPSS_{model} = 1 - \frac{CRPS_{model}}{CRPS_{clim}}.
 \end{equation}
-And using the naive persistence model as a reference, we computed the RMSE Skill Score,
+Similarly, using the naive persistence model as a reference, we computed the RMSE Skill Score,
 \begin{equation}
-RMSE-SS_{model} = 1 - \frac{RMSE_{model}}{RMSE_{naive.}}.
+RMSE-SS_{model} = 1 - \frac{RMSE_{model}}{RMSE_{naive}}.
 \end{equation}
@@
 ## Method Group 1: Statistical Models (Theta)
+
+To compare the performance of neural network models to statistical models, we evaluated forecasts generated by the Theta model.
+Provided a seasonally adjusted univariate time series, the Theta model creates forecasts by modifying the second differences of the data \cite{}.
+The magnitude of local curvature modifications is given by the Theta coefficient,
+
+\begin{equation}
+\nabla^2 Z_t(\theta) = \theta \, \nabla^2 X_t,
+\end{equation}
+
+where $\nabla$ is the difference operator, $Z_t$ is defined as a theta line and $X_t$ is the original data.
+When $\theta < 1$, the second differences are reduced from the data, yielding a Theta line that amplifies the long term trends of the data.
+For instance, if $\theta = 0$, the theta line will be a linear regression of the data.
+And if $\theta > 1$, the data's short term behavior will be magnified in the Theta line.
+The Theta model generates a forecast by extrapolating the linear combination of two or more Theta lines.
+Although the Theta model is relatively simple, it has performed very well in the M3 and M4 forecasting competitions, which has motivated its use in this study as a benchmark statistical model \cite{}.
+In the time since the Theta model was originally presented in 2000, new variations of the Theta model have been presented which have been shown to outperform the original Theta model \cite{}.
+Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing variation of the Theta model.
 

### 2024-05-29 s3_utils.py
Forgot to modify after changing s3_utils

@@ -29,10 +29,10 @@ def read_credentials_from_json(file_path):
         credentials = json.load(f)
     return credentials.get('accessKey'), credentials.get('secretKey')
 
-def upload_df_to_s3(object_name, dataframe, s3_dict={'client': None, 'bucket': None}):
+def upload_df_to_s3(object_name, dataframe, s3_dict={'client': None, 'bucket': None}, index=False):
     # Convert DataFrame to CSV format in memory
     csv_buffer = StringIO()
-    dataframe.to_csv(csv_buffer, index=False)
+    dataframe.to_csv(csv_buffer, index=index)
     
     s3_dict['client'].put_object(
         Bucket=s3_dict['bucket'], 

### 2024-05-29 stats_utils.py

@@ -101,6 +101,6 @@ class AutoThetaForecaster():
                            prediction.time_index[0].strftime('%Y_%m_%d.csv')
             df = prediction.pd_dataframe(suppress_warnings=True)
             if self.s3_dict['client']:
-                upload_df_to_s3(csv_name, df, self.s3_dict)
+                upload_df_to_s3(csv_name, df, self.s3_dict, index=True)
             else:
                 df.to_csv(csv_name)

### 2024-06-11 manuscript/manuscript.Rmd
Headground on results sect

@@
 
-There are many promising neural network architectures that have not been evaluated for water quality forecasting.
-Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network which was published in 1997 \cite{lstm}.
-Over the nearly 30 years since LSTM's were introduced, researchers have identified that LSTM's have notable shortcomings, and there have been continued efforts towards the development of better performing models.
+There are many promising neural network architectures that have not yet been evaluated for water quality forecasting.
+Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network, a model that was published in 1997 \cite{lstm}.
+Over the nearly 30 years since LSTM's were introduced, researchers in computer science have built upon these earlier neural network architectures, introducing new models that achieve state of the art performance \cite{tft, tcn, blockrnn, NBEATS}.
 Many recent studies that use neural networks for water quality forecasting have not explored more contemporary neural network architectures.
@@
 
-We evaluated the forecasting performance of 11 different models on water temperature, dissolved oxygen and chlorophyll-A time series, which were collected by in-situ sensors at 34 freshwater sites across North America.
+We evaluated the forecasting performance of 12 different forecasting models on water temperature, dissolved oxygen and chlorophyll-A time series recorded by in-situ sensors at 34 freshwater sites across North America.
 These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON.
@@
 If the gap size was less than 5 days, then the gap was filled using a Gaussian Process Filter.
-For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, for the case when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.
+For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.
 
@@
 For instance, if $\theta = 0$, the theta line will be a linear regression of the data.
-And if $\theta > 1$, the data's short term behavior will be magnified in the Theta line.
+And if $\theta > 1$, the short term behavior of the data will be magnified in the Theta line.
 The Theta model generates a forecast by extrapolating the linear combination of two or more Theta lines.
-Although the Theta model is relatively simple, it has performed very well in the M3 and M4 forecasting competitions, which has motivated its use in this study as a benchmark statistical model \cite{}.
+Although the Theta model is relatively simple, it has performed remarkably well in prominent forecasting competitions \cite{}.
 In the time since the Theta model was originally presented in 2000, new variations of the Theta model have been presented which have been shown to outperform the original Theta model \cite{}.
-Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing variation of the Theta model.
+Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing model from a selection of variations of the Theta model.
 
@@
 
-For the 8 neural network models that were investigated, there were varying design choices made regarding which type of covariates could be used.
-For instance, models could accept past covariates only, future covariates only or both future and past covariates.
-
-For the models that accept past covariates (**list these models**), we used the other target variables recorded at that site as well as air temperature as covariates.
-For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate.
-The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
-
-We provide fully reproducible code for fitting, scoring and visualizing the forecasts.
-All the machine learning forecasts were generated using the `darts` python library.
-`darts` is similar to other libraries `scikit-learn` in Python or `tidymodels` in R that allow users to easily employ time series forecasting models without having to implement them.
-For model comparison, we rely on the `CRPS` library in Python to efficiently compute the continuous ranked probability scores (CRPS).
-When comparing the probabilistic forecasts, we express the CRPS score in error-orientation -- i.e., a lower CRPS indicates a smaller margin of error.
-
 All the machine learning algorithms that are used in this study are based on neural network architectures.
-Neural networks have the property of being universal function approximators, so theoretically, it is possible that all of these models could exactly approximate the data generating process \cite{}.
+Neural networks have the property of being universal function approximators, so in theory, it is possible that all of these models could exactly approximate the data generating process \cite{}.
 Yet, it is often true that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock, makridakis}.
@@
 We will not go into detail on how these different models work, but for those interested to learn more, we list the models and their references in table \ref{}.
-For readers who do not specialize in ML, an important concept to understand is that the ML models learn directly from the data and are not instilled with domain knowledge.
-This non-mechanisitic basis is at once very powerful as it does not restrict the models with misleading assumptions but is simultaneously limiting as these models are not readily interpretable and generally require more data than mechanistic analogs \cite{}.
-For the management of critical resources like water, this lack of interpretability could be a deterrent to the use of ML methods, and there may not be enough data to accomodate ML approaches \cite{}.
+For readers who do not specialize in ML, an important concept to understand is that the neural network models learn directly from the data and are not instilled with domain knowledge.
+This non-mechanisitic basis is at once very powerful as it does not restrict the models with misleading assumptions, but neural networks are also limiting in that they are not readily interpretable and require more data than knowledge-guided methods \cite{justinread2019}.
+For the management of critical resources like water, the lack of interpretability could be a deterrent to the adoption of NN methods, and there may not be enough data for some water systems to accommodate ML approaches \cite{}.
 
-The ML models are configured to generate probabilistic forecasts.
-The neural network architectures that are used in this study are deterministic, so they are not able to directly represent probability distributions.
+The neural network models are configured to generate probabilistic forecasts.
+The neural network architectures that are used in this study map real-valued vectors to real-valued vectors, so they are not able to directly produce probabilistic forecasts.
 We work around this limitation by performing quantile regression whereby the neural networks are trained to output quantiles at each time step in the forecast window.
@@
 
+For the 8 neural network models that we investigate, there are varying design choices made regarding the type of covariates that can be used.
+For this study, we employed 2 groups of models, one group used past covariates and the other used future covariates.
+For the models that accept past covariates (which included TCN, BlockRNN, NLinear, DLinear, Transformer and NBEATS), we used the other target variables recorded at that site as well as air temperature as covariates.
+For the models that only accept future covariates (which included TFT and LSTM), we used the day of the year as the covariate.
+The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
+
+We provide fully reproducible code for fitting, scoring and visualizing the forecasts.
+All the machine learning forecasts were generated using the `darts` python library \cite{}.
+`darts` is similar to other libraries like `scikit-learn` in Python or `tidymodels` in R in that the library allows users to employ a variety of time series forecasting models without having to implement them.
+
 # Results
+
+We examine the forecast skill of 8 neural network models (LSTM, BlockRNN, NBEATS, NLinear, DLinear, TFT, TCN and Transformer), 1 statistical model (AutoTheta) and 2 null models (Naive Persistence and Climatology) on time series taken from the NEON Forecasting Challenge's Aquatics Theme.
+In addition to these individual models, we created an ensemble model that aggregates the forecasts taken from all the neural network models.
+Following recent work that has established that multi-model ensembles offer advantages over individual models for water temperature forecasting \cite{freya-mme}, we were inspired to investigate multi-model ensembles in this comparative study.
+For the neural network multi-model ensemble, all the NN models are represented equally, hence its name "Naive Ensemble".
+
+In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables.
+For Dissolved Oxygen (DO) and Water Temperature (WT), there are similar results for model performance: the neural network models outperform the AutoTheta, Naive Persistence and Climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE.
+Yet, with Chlorophyll-A, there are much different results: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE.
+BlockRNN attains the best performance with respect to CRPS and the Naive Persistence model attains the best mean RMSE.
+
+Intuition for why we see the results in the scoring tables can be gained by looking at individual forecasts shown in Figure \ref{}.
 
@@
         \centering
-        \caption{Dissolved Oxygen}
+        \caption{Dissolved Oxygen ($\text{mg} \, \text{L}^{-1}$)}
         \begin{tabular}{lcc}
@@
         \centering
-        \caption{Water Temperature}
+        \caption{Water Temperature ($^{\circ} \text{C}$)}
         \begin{tabular}{lcc}
@@
         \centering
-        \caption{Chlorophyll-a}
+        \caption{Chlorophyll-a ($\text{mg} \, \text{L}^{-1}$)}
         \begin{tabular}{lcc}

### 2024-06-12 manuscript/manuscript.Rmd
Making ground on the disc section

@@
 In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables.
-For Dissolved Oxygen (DO) and Water Temperature (WT), there are similar results for model performance: the neural network models outperform the AutoTheta, Naive Persistence and Climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE.
-Yet, with Chlorophyll-A, there are much different results: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE.
-BlockRNN attains the best performance with respect to CRPS and the Naive Persistence model attains the best mean RMSE.
+For Dissolved Oxygen (DO) and Water Temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, Naive Persistence and Climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE.
+Yet, with Chlorophyll-A, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE; and, BlockRNN and the Naive Persistence model attain the best performance with respect to CRPS and RMSE, respectively.
 
-Intuition for why we see the results in the scoring tables can be gained by looking at individual forecasts shown in Figure \ref{}.
+By examining some of the individual forecasts (Figure \ref{}) used for the evaluations, it is possible to gain intuition for why the neural network models perform well across the target variables.
+The AutoTheta model produces forecasts that look similar to a linear regression based upon recently observed values.
+These relatively simple forecasts perform well for many evaluation intervals, but there a few cases when AutoTheta fails significantly, which negatively impacts the model's overall performance.
+For instance, after a peak of dissolved oxygen in the winter of 2024, the AutoTheta model forecasts that DO will continue to increase which is opposed to the trend that DO peaks in the winter and declines through the spring.
+Similarly with Chlorophyll-A, the AutoTheta model wrongly extrapolates that a spike in Chlorophyll-A will lead to even higher Chlorophyll-A concentrations instead of an immediate reversion to the non-bloom state.
+
+Conversely, the neural network models are able to learn from historical trends.
+For instance, at the beginning of 2024, the neural network models have learned from the training data that DO peaks in the winter and declines through the spring, so the neural networks correctly predicts that there will be a decline in dissolved oxygen during this month.
+With Chlorophyll-A, however, we see that the neural network models fail to predict any of the spikes in concentration which originate from algae blooms; instead, the neural networks take a conservative approach, only predicting that the chla concentration will remain at non-bloom levels throughout the year.
+
+These general patterns in model performance can also be seen in the skill score plots displayed in Figures \ref{}.
+The AutoTheta model tends to have longer tails, indicating that the AutoTheta forecasts are more likely to perform very well or very badly.
+Meanwhile, the neural network models generally have fewer forecasts that underperform relative to AutoTheta and less forecasts that are outlier outperformers, but the distributions of NN skill scores are translated towards outperforming.
+In these plots, we display scores according to water body type, and we did not see any significant differences in performance across these categories.
+This may be due to an underreporting of scores for lakes and non-wadeable rivers relative to the number of scores found for wadeable streams.
+
+When examining the performance of the models within a forecast horizon, additional nuances emerge, confusing the perspective that neural network models are the best performing model class.
+In Figure \ref{}, the AutoTheta model is consistently the best performing model for short time horizons across target variables, but by the end of the 30-day horizon, the AutoTheta model is the worst perfoming model universally.
+Meanwhile the neural network models underperform AutoTheta during the early stages of the forecast window (generally, when t < 5), but their forecast skill declines less rapidly than the skill score of AutoTheta.
+So while, the neural networks generally outperform AutoTheta and the null models according to coarse scoring metrics like mean CRPS and RMSE, AutoTheta is the best performer in short horizons.
 
@@
 
-Over the last decade, ML has gained significant prominence after a spate of high profile successes like beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}.
-During this time, researchers have developed many ML models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}.
-For this study, we selected 8 ML models that include s
+Over the last decade, machine learning has gained significant prominence after a spate of high profile successes based on neural network models, some of these successes include beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}.
+During this time, researchers in computer science have developed new neural network models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}.
+Simultaneously during the last 10 years, there has been a growing focus in limnology towards using neural networks for water quality forecasting, but this focus has primarily centered on older neural network architectures and there has been little exploration of newer methods.
+This study aims to address this research gap by performing a comprehensive comparison of neural network models, including models taken from more recent research in computer science.
 
-Yet, these claims of improvement tend to have major limitations, notably that the studies examine a limited amount of time series and they do not compare to naive benchmarks \cite{makridakis}.
+Instead of identifying that there is some new neural network model that performs exceptionally well, this study has affirmed the well-established forecasting axiom that ensemble forecasts are often more accurate and robust than forecasts from single models.
+We found that the Naive Ensemble model which aggregated the forecasts from the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a (attaining the second and third best CRPS and RMSE, respectively).
+This result affirms recent work from \cite{freya} that established multi-model ensembles as advantageous over individual models for water temperature forecasting.
 
-Models tend to struggle to accurately forecast chlorophyll during bloom conditions \cite{woelmer}. Accurately predicting timing and magnitude of bloom events is not surprising
+Yet, as is often the case when judging whether one method is superior to another, we have found that subtle changes to our criteria may have produced radically different results.
+For instance, it is possible that if we changed the forecast horizon to 5 days, then the AutoTheta model may have been the best performing model as the AutoTheta model outperformed all the other models in short time horizons (see Figure \ref{}).
+But, it is important to qualify that changing our criteria to consider shorter term forecasts may result in the neural network models learning different behavior during training.
+How these models perform with different forecast horizons warrants future research.
+
+The neural network models struggled to forecast Chlorophyll-A, a result that is not surprising given that forecasting algae blooms has been well established as a challenging prediction problem in limnology \cite{}.
+The neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chlorophyll-a concentration would remain in a non-bloom state.
+It is possible that the neural network models settled on this conservative behavior as they were not provided enough information to predict blooms -- the neural networks were not provided with some of the typical covariates used in process-based chlorophyll-a models like Nitrogren and Phosphorous concentrations as well as measures of photosynthetic active radiation.
+And, it is likely that this conservative strategy is an artifact of the length of the forecast horizon.
+Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a.
+Predicting the exact time of such an event in a long term horizon will be more difficult than predicting the onset of a bloom in the near-term \cite{}.
+
+We hope that this comparative study will encourage more research that investigates neural networks for water quality forecasting.
+In place of the one research question that we attempted to answer, "How do neural network models perform for water quality forecasting?", many new questions and ideas have arose.
+Outside of exploring how the performance of neural networks will vary with forecast horizon, which was mentionned previously, we think that some other promising directions for exploration include cross-learning and hybrid models.
+
+Provided that there are large limnological datasets that currently exist, and there will be an increasing amount of limnological data coming online, neural network models offer significant advantages over process-based and statistical models as they will be able to learn directly from the data.
+This data-driven approach is advantageous in a few ways: as NN models can learn directly from the data, these models will not be limited by the assumptions inherent to process-based and statistical models.
+Moreover, these models should improve as
 

### 2024-06-13 manuscript/manuscript.Rmd
Additions on disc

@@
 Simultaneously during the last 10 years, there has been a growing focus in limnology towards using neural networks for water quality forecasting, but this focus has primarily centered on older neural network architectures and there has been little exploration of newer methods.
-This study aims to address this research gap by performing a comprehensive comparison of neural network models, including models taken from more recent research in computer science.
+This study aims to address this research gap by performing a comprehensive comparison of neural network models, including models taken from more recent developments in computer science.
 
 Instead of identifying that there is some new neural network model that performs exceptionally well, this study has affirmed the well-established forecasting axiom that ensemble forecasts are often more accurate and robust than forecasts from single models.
-We found that the Naive Ensemble model which aggregated the forecasts from the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a (attaining the second and third best CRPS and RMSE, respectively).
-This result affirms recent work from \cite{freya} that established multi-model ensembles as advantageous over individual models for water temperature forecasting.
+We found that an ensemble model which aggregates the forecasts from the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a, attaining the second and third best CRPS and RMSE, respectively.
+This result affirms recent work from \cite{freya} which established that multi-model ensembles offer advantages over individual models for water temperature forecasting.
 
 Yet, as is often the case when judging whether one method is superior to another, we have found that subtle changes to our criteria may have produced radically different results.
-For instance, it is possible that if we changed the forecast horizon to 5 days, then the AutoTheta model may have been the best performing model as the AutoTheta model outperformed all the other models in short time horizons (see Figure \ref{}).
-But, it is important to qualify that changing our criteria to consider shorter term forecasts may result in the neural network models learning different behavior during training.
-How these models perform with different forecast horizons warrants future research.
+For instance, if did not change any of the model training configurations but cut off the forecast horizon to 5 days, then the AutoTheta model would have been the best performing model since the AutoTheta model outperformed all the other models in short time horizons (see Figure \ref{}).
+But, it is important to qualify that if we changed the model training configurations to consider shorter forecasts horizons, this could result in the neural network models learning different behavior during training.
+How neural networks models perform with different forecast horizons warrants future research.
 
-The neural network models struggled to forecast Chlorophyll-A, a result that is not surprising given that forecasting algae blooms has been well established as a challenging prediction problem in limnology \cite{}.
-The neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chlorophyll-a concentration would remain in a non-bloom state.
-It is possible that the neural network models settled on this conservative behavior as they were not provided enough information to predict blooms -- the neural networks were not provided with some of the typical covariates used in process-based chlorophyll-a models like Nitrogren and Phosphorous concentrations as well as measures of photosynthetic active radiation.
-And, it is likely that this conservative strategy is an artifact of the length of the forecast horizon.
+The neural network models struggled to forecast Chlorophyll-A, a result that is not surprising given that forecasting algae blooms is well established as a challenging prediction problem in limnology \cite{}.
+With Cholorophyll-A (chla), the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state.
+It is possible that the neural network models settled on this conservative behavior because we did provide enough information to predict blooms; the neural networks were not provided with some of the typical covariates used in process-based models like Nitrogen and Phosphorous concentrations, and photosynthetic active radiation.
+It is also likely that this conservative strategy is an artifact of the length of the forecast horizon.
 Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a.
 Predicting the exact time of such an event in a long term horizon will be more difficult than predicting the onset of a bloom in the near-term \cite{}.
+It seems reasonable that the neural networks would take a conservative approach to forecasting chla as making the wrong prediction for an algae bloom will be costly with respect to the model's objective.
 
-We hope that this comparative study will encourage more research that investigates neural networks for water quality forecasting.
-In place of the one research question that we attempted to answer, "How do neural network models perform for water quality forecasting?", many new questions and ideas have arose.
-Outside of exploring how the performance of neural networks will vary with forecast horizon, which was mentionned previously, we think that some other promising directions for exploration include cross-learning and hybrid models.
+In place of the one research question that we set out to answer, "How do neural network models perform for water quality forecasting?", many new questions and ideas have arose from this project.
+Outside of exploring how the performance of neural networks will vary with forecast horizon, we think that some other promising directions for exploration include cross-learning and hybrid models.
+Throughout this study, we have trained models individually for each target time series; cross-learning presents a different approach where we would train one model across all the time series available for a given target variable.
+Through cross-learning, the model might be able to learn a pattern from one time series that will improve forecasts on a different time series; but, if the model is trained on a single time series, then the model may never learn this pattern.
+We also reckon that hybrid models may have superior forecasting performance to neural network or statistical models individually.
+We observed that AutoTheta were the best performing models early in the forecast horizon, and neural networks performed best later in the forecast horizon.
+This motivates the exploration of a hybrid model that employs AutoTheta at the beginning of the forecast horizon and neural network models later in the forecast horizon.
 
-Provided that there are large limnological datasets that currently exist, and there will be an increasing amount of limnological data coming online, neural network models offer significant advantages over process-based and statistical models as they will be able to learn directly from the data.
-This data-driven approach is advantageous in a few ways: as NN models can learn directly from the data, these models will not be limited by the assumptions inherent to process-based and statistical models.
-Moreover, these models should improve as
+It is an exciting time to study water quality forecasting.
+There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as more sensors will be deployed \cite{}.
+Neural network models offer promising advantages over alternative methods for limnological forecasting as these methods can learn complex patterns without being restrained by a need for domain knowledge.
+Furthermore, their performance has the potential to improve as more water quality data becomes available.
+Yet, they also present a range of problems like a lack of interpretability and generalizability that could restrain them from being useful decision support tools for the safety critical problem of water resource management.
+This study does not attempt to provide a definitive answer as to whether neural networks will be the best method for water quality forecasting as answering this questions will take numerous studies to reach a consensus on.
+Instead, we hope that this study will generate ideas and a collective momentum towards improving our general ability to forecast and manage water quality.
 

### 2024-06-13 manuscript/manuscript.Rmd
Captions done

@@
 Freshwater ecosystems have been disproportionately impacted by anthropogenic activities, a trend that is expected to continue in the future \cite{albert2021}.
-Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{}.
+Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, influence many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{}.
 Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{Hanson, zwart}.
@@
 For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.
+The intuition behind this method for missing data impuation is that the neural network models will be biased towards the model that you use to fill in missing data.
+Since it has been established that a climatology model is a top performing model throughout the NEON forecasting challenge, we reckoned that biasing the neural network models towards the climatology model will likely bias the models towards making better predictions.
 
@@
 
-We examine the forecast skill of 8 neural network models (LSTM, BlockRNN, NBEATS, NLinear, DLinear, TFT, TCN and Transformer), 1 statistical model (AutoTheta) and 2 null models (Naive Persistence and Climatology) on time series taken from the NEON Forecasting Challenge's Aquatics Theme.
+We examine the forecast skill of 8 neural network models (LSTM, BlockRNN, NBEATS, NLinear, DLinear, TFT, TCN and Transformer), 1 statistical model (AutoTheta) and 2 null models (naive persistence and climatology) on time series taken from the NEON Forecasting Challenge's Aquatics Theme.
 In addition to these individual models, we created an ensemble model that aggregates the forecasts taken from all the neural network models.
@@
 In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables.
-For Dissolved Oxygen (DO) and Water Temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, Naive Persistence and Climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE.
-Yet, with Chlorophyll-A, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE; and, BlockRNN and the Naive Persistence model attain the best performance with respect to CRPS and RMSE, respectively.
+For dissolved oxygen (DO) and water temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, naive persistence and climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE.
+Yet, with Chlorophyll-A, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE; and, BlockRNN and the naive persistence model attain the best performance with respect to CRPS and RMSE, respectively.
 
@@
     \centering
-    \caption{Mean CRPS and RMSE}
     \begin{subtable}{\textwidth}
@@
     \end{subtable}
+    \caption{Mean CRPS and RMSE for dissolved oxygen, water temperature and chlorophyll-a forecasts.
+The neural network-based models generally outperform the statistical benchmark model, AutoTheta, as well as the null models, naive persistence and climatology, across the target variables.
+For dissolved oxygen and water temperature, the NN ensemble model is the best performing model with respect to CRPS and RMSE.
+With chlorophyll-a, BlockRNN attains the best CRPS score, and naive persistence attains the best RMSE.}
     \label{tab:mean_crps_rmse}
@@
   \end{subfigure}
-  \caption{Comparison of Historical and Naive Oxygen}
+  \caption{Skill score distributions for dissolved oxygen.
+The CRPSS plot measures the CRPS skill relative to the climatology model.
+The RMSE-SS plot measures the RMSE skill relative to the naive persistence model.
+A guideline is plotted at the threshold for underperformance: a score above this line denotes that a forecast is underperforming the reference model; a score below this line denotes that a forecast is outperforming the reference model.
+The AutoTheta model exhibits longer tails relative to most of the neural network models, indicating that AutoTheta has a larger amount of forecasts that do either very well or very poorly.
+The neural network models tend to have shorter tails in the underperforming region.}
 \end{figure}
@@
   \end{subfigure}
-  \caption{Comparison of Historical and Naive temperature}
+  \caption{Skill score distributions for water temperature.
+AutoTheta does not have the long tails as seen in the skill score distributions for dissolved oxygen and chlorophyll-a, instead the neural network models have heavier tails in the outperforming region.
+This indicates that the neural network models are producing high accuracy forecasts more frequently than the AutoTheta model.}
 \end{figure}
@@
   \end{subfigure}
-  \caption{Comparison of Historical and Naive chla}
+  \caption{Skill score distributions for chlorophyll-a.
+The neural network models have shorter tails in the underperforming region relative to AutoTheta, indicating that the neural network models are producing a smaller amount of inaccurate forecasts.
+The naive persistence null model attained the lowest RMSE on chlorophyll-a, and this is apparent in the RMSE-SS plot as the models have skill score distributions that are generally centered above the underperformance threshold.}
 \end{figure}
@@
 
-  \caption{Oxygen, Temperature, Chlorophyll Intrawindow plots}
+  \caption{Mean skill scores within the 30-day forecast horizon.
+In the early phase of the forecast horizon (t < 5), the AutoTheta model is universally the top performing model.
+However, the performance of the AutoTheta model rapidly declines over the course of the 30-day horizon, concluding in AutoTheta being the worst performing model universally at the end of the horizon.
+The neural network models do not perform as well as AutoTheta at the beginning of the forecast horizon, but their performance declines less rapidly throughout the horizon.}
 \end{figure}
@@
   \end{subfigure}
-  \caption{Individual Forecast Examples}
+  \caption{Individual forecasts of dissolved oxygen and chlorophyll-a.
+The AutoTheta model produces accurate forecasts most of the year, especially for dissolved oxygen; but, there are some intervals where AutoTheta produces extremely inaccurate forecasts.
+After peaks in dissolved oxygen and chlorophyll-a, the AutoTheta model wrongly predicts that the values will continue to increase.
+Conversely, the neural network-based forecasts from the TFT and the Naive Ensemble model have learned from the training data that high levels of dissolved oxygen in the winter are followed by a steady decline throughout the spring.
+With chlorophyll-a, the neural networks take a conservative strategy where they predict that chlorophyll-a will remain in a non-bloom state, and these models never take a risk in forecasting that a bloom will occur.}
 \end{figure}

### 2024-06-13 manuscript/manuscript.Rmd
Adding pngs to manuscript dir

@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/sitetype_legend_0.png}
+    \includegraphics[width=\textwidth]{./plots/sitetype_legend_0.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/type_local_historical_oxygen.png}
+    \includegraphics[width=\textwidth]{./plots/type_local_historical_oxygen.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/type_local_naive_oxygen.png}
+    \includegraphics[width=\textwidth]{./plots/type_local_naive_oxygen.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/type_local_historical_temperature.png}
+    \includegraphics[width=\textwidth]{./plots/type_local_historical_temperature.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/type_local_naive_temperature.png}
+    \includegraphics[width=\textwidth]{./plots/type_local_naive_temperature.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/sitetype_legend_1.png}
+    \includegraphics[width=\textwidth]{./plots/sitetype_legend_1.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/type_local_historical_chla.png}
+    \includegraphics[width=\textwidth]{./plots/type_local_historical_chla.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/type_local_naive_chla.png}
+    \includegraphics[width=\textwidth]{./plots/type_local_naive_chla.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/intra_historical_oxygen.png}
+    \includegraphics[width=\textwidth]{./plots/intra_historical_oxygen.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/intra_naive_oxygen.png}
+    \includegraphics[width=\textwidth]{./plots/intra_naive_oxygen.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/intra_historical_temperature.png}
+    \includegraphics[width=\textwidth]{./plots/intra_historical_temperature.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/intra_naive_temperature.png}
+    \includegraphics[width=\textwidth]{./plots/intra_naive_temperature.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/intra_legend.png}
+    \includegraphics[width=\textwidth]{./plots/intra_legend.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/intra_historical_chla.png}
+    \includegraphics[width=\textwidth]{./plots/intra_historical_chla.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/intra_naive_chla.png}
+    \includegraphics[width=\textwidth]{./plots/intra_naive_chla.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/forecast_legend.png}
+    \includegraphics[width=\textwidth]{./plots/forecast_legend.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/oxygen_theta.png}
+    \includegraphics[width=\textwidth]{./plots/oxygen_theta.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/chla_theta.png}
+    \includegraphics[width=\textwidth]{./plots/chla_theta.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/oxygen_tft.png}
+    \includegraphics[width=\textwidth]{./plots/oxygen_tft.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/chla_tft.png}
+    \includegraphics[width=\textwidth]{./plots/chla_tft.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/oxygen_ensemble.png}
+    \includegraphics[width=\textwidth]{./plots/oxygen_ensemble.png}
   \end{subfigure}
@@
     \centering
-    \includegraphics[width=\textwidth]{../plots/chla_ensemble.png}
+    \includegraphics[width=\textwidth]{./plots/chla_ensemble.png}
   \end{subfigure}

### 2024-06-19 manuscript/manuscript.Rmd
Getting a few refs up

@@
 Freshwater ecosystems have been disproportionately impacted by anthropogenic activities, a trend that is expected to continue in the future \cite{albert2021}.
-Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, influence many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{}.
-Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{Hanson, zwart}.
+Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, influence many biological and physical processes in freshwater ecosystems, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{ouellet-proulx, stajkowski, chen_combining}.
+Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{hanson, ladwig}.
 A shortcoming of past limnological forecasting studies that support the use of machine learning is that they tend to focus on a selection of sites that are not representative of freshwater systems across a broad geographic scale.

### 2024-06-21 manuscript/manuscript.Rmd
More ref scanning

@@
 Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, influence many biological and physical processes in freshwater ecosystems, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{ouellet-proulx, stajkowski, chen_combining}.
-Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{hanson, ladwig}.
+Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{hanson, zwart2023}.
 A shortcoming of past limnological forecasting studies that support the use of machine learning is that they tend to focus on a selection of sites that are not representative of freshwater systems across a broad geographic scale.
@@
 
-The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{Quinn}.
-A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}.
-For instance, in a model comparison that examined the forecasts for phenological time series, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
+The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{thomas_span_2023}.
+A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{wheeler_predicting_2024, thomas_nearterm_2023}.
+For instance, in a model comparison that examined the forecasts for phenological time series, \cite{wheeler_predicting_2024} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
 Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the historical null model.
@@
 There are many promising neural network architectures that have not yet been evaluated for water quality forecasting.
-Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network, a model that was published in 1997 \cite{lstm}.
-Over the nearly 30 years since LSTM's were introduced, researchers in computer science have built upon these earlier neural network architectures, introducing new models that achieve state of the art performance \cite{tft, tcn, blockrnn, NBEATS}.
+Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network, a model that was published in 1997 \cite{hochreiter_long_1997}.
+Over the nearly 30 years since LSTM's were introduced, researchers in computer science have built upon these earlier neural network architectures, introducing new models that achieve state of the art performance \cite{lim_temporal_2019, bai_empirical_2018, oreshkin_n-beats_2019}.
 Many recent studies that use neural networks for water quality forecasting have not explored more contemporary neural network architectures.
@@
 For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.
-The intuition behind this method for missing data impuation is that the neural network models will be biased towards the model that you use to fill in missing data.
-Since it has been established that a climatology model is a top performing model throughout the NEON forecasting challenge, we reckoned that biasing the neural network models towards the climatology model will likely bias the models towards making better predictions.
+The intuition behind this method for missing data imputation is that the neural network models will be biased towards the model that you use to fill in missing data.
+Since it has been established that a climatology model is a top performing model throughout the NEON forecasting challenge, we reckoned that biasing the neural network models towards the climatology model will likely bias the models towards making better predictions \cite{wheeler_predicting_2024, thomas_nearterm_2023}.
 
 We compared the performance of the neural network models to the climatology and naive persistence model.
-The climatology model generates forecasts by finding the daily mean and standard deviation and draws samples from a Gaussian Distribution with these parameters.
+The climatology model generates forecasts by finding the daily mean and standard deviation and draws samples from a Gaussian distribution with these parameters.
 The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast horizon.
@@
 To compare the performance of neural network models to statistical models, we evaluated forecasts generated by the Theta model.
-Provided a seasonally adjusted univariate time series, the Theta model creates forecasts by modifying the second differences of the data \cite{}.
+Provided a seasonally adjusted univariate time series, the Theta model creates forecasts by modifying the second differences of the data \cite{assimakopoulos_theta_2000}.
 The magnitude of local curvature modifications is given by the Theta coefficient,
@@
 The Theta model generates a forecast by extrapolating the linear combination of two or more Theta lines.
-Although the Theta model is relatively simple, it has performed remarkably well in prominent forecasting competitions \cite{}.
-In the time since the Theta model was originally presented in 2000, new variations of the Theta model have been presented which have been shown to outperform the original Theta model \cite{}.
+Although the Theta model is relatively simple, it has performed remarkably well in prominent forecasting competitions \cite{makridakis_m3-competition_2000, makridakis_m4_2020}.
+In the time since the Theta model was originally presented in 2000, new variations of the Theta model have been presented which have been shown to outperform the original Theta model \cite{fiorucci_models_2016}.
 Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing model from a selection of variations of the Theta model.
@@
 All the machine learning algorithms that are used in this study are based on neural network architectures.
-Neural networks have the property of being universal function approximators, so in theory, it is possible that all of these models could exactly approximate the data generating process \cite{}.
-Yet, it is often true that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock, makridakis}.
-This performance gap can be due to a variety of reasons including overfitting and insufficient hyperparameter tuning \cite{adcock}.
+Neural networks have the property of being universal function approximators, so in theory, it is possible that all of these models could exactly approximate the data generating process \cite{hanin_universal_2019}.
+Yet, it is often true that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock_gap_2020}.
+This performance gap can be due to a variety of reasons including overfitting and insufficient hyperparameter tuning \cite{adcock_gap_2020}.
 The discrepancy between theory and practice in function approximation with neural networks motivates research on how machine learning models perform in specific domains.

### 2024-06-27 manuscript/manuscript.Rmd
Another edit pass

@@
 The ability to accurately forecast water quality variables has become increasingly important in the era of global change.
-Freshwater ecosystems have been disproportionately impacted by anthropogenic activities, a trend that is expected to continue in the future \cite{albert2021}.
-Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, influence many biological and physical processes in freshwater ecosystems, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{ouellet-proulx, stajkowski, chen_combining}.
+Freshwater ecosystems have been disproportionately impacted by anthropogenic activities, a trend that is expected to continue throughout the 21st century \cite{albert2021}.
+Since water quality variables, like water temperature, dissolved oxygen and chlorophyll-a concentration, influence many biological and physical processes in freshwater ecosystems, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{ouellet-proulx, stajkowski, chen_combining}.
 Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{hanson, zwart2023}.
-A shortcoming of past limnological forecasting studies that support the use of machine learning is that they tend to focus on a selection of sites that are not representative of freshwater systems across a broad geographic scale.
+A shortcoming of past limnological forecasting studies that support the use of machine learning is that they tend to focus on a narrow selection of sites that are not representative of freshwater systems across a broad geographic scale.
 This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by evaluating forecasts at 34 different sites across North America at different times of the year.
 
-The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{thomas_span_2023}.
+The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, a competition where teams can submit forecasts for data that is collected and made publicly accessible by NEON \cite{thomas_span_2023}.
 A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{wheeler_predicting_2024, thomas_nearterm_2023}.
-For instance, in a model comparison that examined the forecasts for phenological time series, \cite{wheeler_predicting_2024} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
+For instance, in a model comparison that examined the forecasts for phenology, \cite{wheeler_predicting_2024} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model.
 Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the historical null model.
@@
 Many recent studies that use neural networks for water quality forecasting have not explored more contemporary neural network architectures.
-We will address this gap in the literature by comparing 8 neural network models including LSTM's as well as more recently published neural network models.
+We will address this gap in the literature by comparing 8 neural network models including LSTM's as well as more recently developed neural network models.
 
@@
 
-We evaluated the forecasting performance of 12 different forecasting models on water temperature, dissolved oxygen and chlorophyll-A time series recorded by in-situ sensors at 34 freshwater sites across North America.
+We evaluated the forecasting performance of 12 different forecasting models on water temperature, dissolved oxygen and chlorophyll-a time series recorded by in-situ sensors at 34 freshwater sites across North America.
 These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON.
@@
 Additionally, maintenance issues led to gaps in the data which also varied across locations.
-Provided the lack of time series and large gaps at certain sites, we evaluated the forecast performance for water temperature, dissolved oxygen and chlorophyll-A at 33, 32 and 9 sites, respectively.
+Provided the lack of time series and large gaps at certain sites, we evaluated the forecast performance for water temperature, dissolved oxygen and chlorophyll-A at 33, 32 and 9 sites respectively.
 
@@
 For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.
-The intuition behind this method for missing data imputation is that the neural network models will be biased towards the model that you use to fill in missing data.
-Since it has been established that a climatology model is a top performing model throughout the NEON forecasting challenge, we reckoned that biasing the neural network models towards the climatology model will likely bias the models towards making better predictions \cite{wheeler_predicting_2024, thomas_nearterm_2023}.
+The intuition behind this method for missing data imputation is that this method will bias the neural networks towards the climatology model.
+Since it has been established that a climatology model produces top performing forecasts throughout the NEON forecasting challenge, backfilling with the climatology model seems likely to induce improved accuracy for the neural network models \cite{wheeler_predicting_2024, thomas_nearterm_2023}.
 
@@
 The climatology model generates forecasts by finding the daily mean and standard deviation and draws samples from a Gaussian distribution with these parameters.
-The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast horizon.
+The naive persistence model finds the last observed value from the target time series and predicts this value for each day in the forecast horizon.
 For each model, we computed the Continuous Ranked Probability Score (CRPS) and the Root Mean Square Error (RMSE).
@@
 
-To compare the performance of neural network models to statistical models, we evaluated forecasts generated by the Theta model.
+To compare the performance of neural network models with a representative statistical model, we evaluated forecasts generated by the Theta model.
 Provided a seasonally adjusted univariate time series, the Theta model creates forecasts by modifying the second differences of the data \cite{assimakopoulos_theta_2000}.
@@
 The Theta model generates a forecast by extrapolating the linear combination of two or more Theta lines.
-Although the Theta model is relatively simple, it has performed remarkably well in prominent forecasting competitions \cite{makridakis_m3-competition_2000, makridakis_m4_2020}.
-In the time since the Theta model was originally presented in 2000, new variations of the Theta model have been presented which have been shown to outperform the original Theta model \cite{fiorucci_models_2016}.
-Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing model from a selection of variations of the Theta model.
+Although the Theta model is relatively simple, it has performed remarkably well in prominent forecasting competitions that include hard to predict data \cite{makridakis_m3-competition_2000, makridakis_m4_2020}.
+Since the Theta model was originally presented in 2000, new variations of the Theta model have been developed that outperform the original Theta model \cite{fiorucci_models_2016}.
+Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing model from a range of Theta model variants.
 
@@
 All the machine learning algorithms that are used in this study are based on neural network architectures.
-Neural networks have the property of being universal function approximators, so in theory, it is possible that all of these models could exactly approximate the data generating process \cite{hanin_universal_2019}.
-Yet, it is often true that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock_gap_2020}.
+Neural networks have the property of being universal function approximators, so it is possible, in theory, that all of these models could exactly approximate the data generating process \cite{hanin_universal_2019}.
+Yet, it is often the case that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock_gap_2020}.
 This performance gap can be due to a variety of reasons including overfitting and insufficient hyperparameter tuning \cite{adcock_gap_2020}.
@@
 We will not go into detail on how these different models work, but for those interested to learn more, we list the models and their references in table \ref{}.
-For readers who do not specialize in ML, an important concept to understand is that the neural network models learn directly from the data and are not instilled with domain knowledge.
-This non-mechanisitic basis is at once very powerful as it does not restrict the models with misleading assumptions, but neural networks are also limiting in that they are not readily interpretable and require more data than knowledge-guided methods \cite{justinread2019}.
-For the management of critical resources like water, the lack of interpretability could be a deterrent to the adoption of NN methods, and there may not be enough data for some water systems to accommodate ML approaches \cite{}.
+An important concept to understand is that the neural network models considered in this study learn directly from the data and are not instilled with domain knowledge.
+This non-mechanisitic basis is at once very powerful as it does not restrict the models with misleading assumptions, but neural networks are also limiting in that they are not readily interpretable and often require more data than knowledge-guided or empirical methods to perform well \cite{karpatne_knowledge-guided_2024, read_processguided_2019}.
+For the management of critical resources like water, the lack of interpretability could be a deterrent to the adoption of NN methods, and there may not be enough data for some water systems to accommodate ML approaches \cite{zhi_deep_2024}.
 
-The neural network models are configured to generate probabilistic forecasts.
-The neural network architectures that are used in this study map real-valued vectors to real-valued vectors, so they are not able to directly produce probabilistic forecasts.
+The neural network models are configured in this study to provide estimates for predictive uncertainty.
+The neural network architectures that we use are deterministic, so they are not able to produce probabilistic forecasts intrinsically.
 We work around this limitation by performing quantile regression whereby the neural networks are trained to output quantiles at each time step in the forecast window.
-Forecasts are then generated by drawing samples according to these quantiles.
+Probabilistic forecasts are then generated by drawing samples according to these quantiles.
 
 For the 8 neural network models that we investigate, there are varying design choices made regarding the type of covariates that can be used.
-For this study, we employed 2 groups of models, one group used past covariates and the other used future covariates.
+For this study, we employed 2 groups of models: one group used past covariates and the other used future covariates.
 For the models that accept past covariates (which included TCN, BlockRNN, NLinear, DLinear, Transformer and NBEATS), we used the other target variables recorded at that site as well as air temperature as covariates.
-For the models that only accept future covariates (which included TFT and LSTM), we used the day of the year as the covariate.
-The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
+For the models that only accept future covariates (which included TFT and RNN), we used the day of the year as the sole covariate.
+All the time series were split into training and validation sets, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.
 
 We provide fully reproducible code for fitting, scoring and visualizing the forecasts.
-All the machine learning forecasts were generated using the `darts` python library \cite{}.
+All the machine learning forecasts were generated using the `darts` python library \cite{herzen_darts_2021}.
 `darts` is similar to other libraries like `scikit-learn` in Python or `tidymodels` in R in that the library allows users to employ a variety of time series forecasting models without having to implement them.
+The github repo for this study can be found at https://github.com/boettiger-lab/neon4cast-darts-ml.
 
@@
 
-We examine the forecast skill of 8 neural network models (LSTM, BlockRNN, NBEATS, NLinear, DLinear, TFT, TCN and Transformer), 1 statistical model (AutoTheta) and 2 null models (naive persistence and climatology) on time series taken from the NEON Forecasting Challenge's Aquatics Theme.
-In addition to these individual models, we created an ensemble model that aggregates the forecasts taken from all the neural network models.
-Following recent work that has established that multi-model ensembles offer advantages over individual models for water temperature forecasting \cite{freya-mme}, we were inspired to investigate multi-model ensembles in this comparative study.
-For the neural network multi-model ensemble, all the NN models are represented equally, hence its name "Naive Ensemble".
+We examine the forecast skill of 8 neural network models (RNN, BlockRNN, NBEATS, NLinear, DLinear, TFT, TCN and Transformer), 1 statistical model (AutoTheta) and 2 null models (naive persistence and climatology) on time series taken from the NEON Forecasting Challenge's Aquatics Theme.
+Following recent work that has established that multi-model ensembles offer advantages over individual models for water temperature forecasting \cite{olsson_multimodel_2024}, we were inspired to investigate multi-model ensembles in this comparative study.
+So, in addition to the ML, empirical and null models, we created an ensemble model that aggregates the forecasts taken from all the neural network models.
+For this neural network ensemble, all the NN models are represented equally, hence its name "Naive Ensemble".
 
@@
 For dissolved oxygen (DO) and water temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, naive persistence and climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE.
-Yet, with Chlorophyll-A, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE; and, BlockRNN and the naive persistence model attain the best performance with respect to CRPS and RMSE, respectively.
+Yet, with chlorophyll-a, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE.
+Instead, BlockRNN and the naive persistence model attain the best performance with respect to CRPS and RMSE, respectively.
 
-By examining some of the individual forecasts (Figure \ref{}) used for the evaluations, it is possible to gain intuition for why the neural network models perform well across the target variables.
-The AutoTheta model produces forecasts that look similar to a linear regression based upon recently observed values.
-These relatively simple forecasts perform well for many evaluation intervals, but there a few cases when AutoTheta fails significantly, which negatively impacts the model's overall performance.
+By examining some of the individual forecasts, as shown in Figure \ref{}, it is possible to gain intuition for why the neural network models perform well across the target variables.
+The AutoTheta model produces forecasts that resemble linear regressions based upon recently observed values.
+These relatively simple forecasts perform well for many evaluation intervals, but there a few cases when AutoTheta fails catastrophically, negatively impacting the model's overall performance.
 For instance, after a peak of dissolved oxygen in the winter of 2024, the AutoTheta model forecasts that DO will continue to increase which is opposed to the trend that DO peaks in the winter and declines through the spring.
-Similarly with Chlorophyll-A, the AutoTheta model wrongly extrapolates that a spike in Chlorophyll-A will lead to even higher Chlorophyll-A concentrations instead of an immediate reversion to the non-bloom state.
+Similarly with Chlorophyll-A, the AutoTheta model wrongly extrapolates that a spike in chlorophyll-a will lead to higher chlorophyll-a concentrations instead of an immediate reversion to the non-bloom state.
 
 Conversely, the neural network models are able to learn from historical trends.
-For instance, at the beginning of 2024, the neural network models have learned from the training data that DO peaks in the winter and declines through the spring, so the neural networks correctly predicts that there will be a decline in dissolved oxygen during this month.
-With Chlorophyll-A, however, we see that the neural network models fail to predict any of the spikes in concentration which originate from algae blooms; instead, the neural networks take a conservative approach, only predicting that the chla concentration will remain at non-bloom levels throughout the year.
+For instance, at the beginning of 2024, the neural network models have learned from the training data that DO peaks in the winter and declines through the spring.
+With chlorophyll-a, however, we see that the neural network models fail to predict any of the spikes in concentration which originate from algae blooms; instead, the neural networks take a conservative approach, only predicting that the chla concentration will remain at non-bloom levels throughout the year.
 
-These general patterns in model performance can also be seen in the skill score plots displayed in Figures \ref{}.
+These general patterns in model performance can be observed in the skill score plots displayed in Figures \ref{}.
 The AutoTheta model tends to have longer tails, indicating that the AutoTheta forecasts are more likely to perform very well or very badly.
-Meanwhile, the neural network models generally have fewer forecasts that underperform relative to AutoTheta and less forecasts that are outlier outperformers, but the distributions of NN skill scores are translated towards outperforming.
-In these plots, we display scores according to water body type, and we did not see any significant differences in performance across these categories.
-This may be due to an underreporting of scores for lakes and non-wadeable rivers relative to the number of scores found for wadeable streams.
+Meanwhile, the neural network models generally have fewer forecasts that underperform relative to AutoTheta as well as fewer forecasts that are outlier outperformers.
+The distributions of NN skill scores are more skewed towards outperforming than AutoTheta's skill scores are.
+In these plots, we display scores according to water body type, but we did not see any significant differences in performance across these categories.
+However, this may be due to an underreporting of scores for lakes and non-wadeable rivers relative to the number of scores found for wadeable streams.
 
@@
 
-Over the last decade, machine learning has gained significant prominence after a spate of high profile successes based on neural network models, some of these successes include beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}.
-During this time, researchers in computer science have developed new neural network models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}.
-Simultaneously during the last 10 years, there has been a growing focus in limnology towards using neural networks for water quality forecasting, but this focus has primarily centered on older neural network architectures and there has been little exploration of newer methods.
-This study aims to address this research gap by performing a comprehensive comparison of neural network models, including models taken from more recent developments in computer science.
+Over the last decade, machine learning has gained significant prominence after a spate of high profile successes based on neural network models, some of these successes include beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{silver_mastering_2017, jumper_highly_2021, zhao_survey_2023}.
+During this time, researchers in computer science have developed new neural network models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis_m4_2020}.
+Simultaneously during the last 10 years, there has been a growing focus in limnology towards using neural networks for water quality forecasting, but this focus has primarily centered on older neural network architectures witn little exploration of newer methods.
+This study aims to address this research gap by performing a comprehensive comparison of neural network models that include more recently developed models.
 
-Instead of identifying that there is some new neural network model that performs exceptionally well, this study has affirmed the well-established forecasting axiom that ensemble forecasts are often more accurate and robust than forecasts from single models.
-We found that an ensemble model which aggregates the forecasts from the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a, attaining the second and third best CRPS and RMSE, respectively.
-This result affirms recent work from \cite{freya} which established that multi-model ensembles offer advantages over individual models for water temperature forecasting.
+Instead of identifying that there is a new neural network model that performs exceptionally well, this study has affirmed the conventional wisdom that ensemble forecasts are often more accurate and robust than forecasts from individual models.
+We found that an ensemble model which aggregates the forecasts from all the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a, attaining the second and third best CRPS and RMSE respectively.
+This result affirms recent work from \cite{olsson_multimodel_2024} which established that multi-model ensembles offer advantages over individual models for water temperature forecasting.
 
 Yet, as is often the case when judging whether one method is superior to another, we have found that subtle changes to our criteria may have produced radically different results.
-For instance, if did not change any of the model training configurations but cut off the forecast horizon to 5 days, then the AutoTheta model would have been the best performing model since the AutoTheta model outperformed all the other models in short time horizons (see Figure \ref{}).
-But, it is important to qualify that if we changed the model training configurations to consider shorter forecasts horizons, this could result in the neural network models learning different behavior during training.
-How neural networks models perform with different forecast horizons warrants future research.
+For instance, if we had use the same forecasts abbreviated to a 5 day horizon, then the AutoTheta model would have been the best performing model according to the scoring metrics used in this study.
+But, it is important to qualify that if we changed the model training configurations to consider shorter forecasts horizons, this would produce forecasts that will be different than the ones we evaluated for the 30-day horizon.
+Thus, how neural networks models perform over different forecast horizons warrants future research.
 
-The neural network models struggled to forecast Chlorophyll-A, a result that is not surprising given that forecasting algae blooms is well established as a challenging prediction problem in limnology \cite{}.
-With Cholorophyll-A (chla), the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state.
+The neural network models struggled to forecast chlorophyll-a, a result that is not surprising given that forecasting algae blooms is well established as a challenging prediction problem in limnology \cite{chen_combining}.
+With cholorophyll-a, the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state.
 It is possible that the neural network models settled on this conservative behavior because we did provide enough information to predict blooms; the neural networks were not provided with some of the typical covariates used in process-based models like Nitrogen and Phosphorous concentrations, and photosynthetic active radiation.
-It is also likely that this conservative strategy is an artifact of the length of the forecast horizon.
+Yet, it is also plausible that this conservative strategy was driven by the length of the forecast horizon.
 Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a.
-Predicting the exact time of such an event in a long term horizon will be more difficult than predicting the onset of a bloom in the near-term \cite{}.
-It seems reasonable that the neural networks would take a conservative approach to forecasting chla as making the wrong prediction for an algae bloom will be costly with respect to the model's objective.
+Predicting the timing of such an event in a long term horizon will be more difficult than performing the same prediction on a shorter horizon as there are more opportunities in the long term horizon to be wrong.
+With this reasoning, it is sensible that the neural networks would take a conservative approach.
 
-In place of the one research question that we set out to answer, "How do neural network models perform for water quality forecasting?", many new questions and ideas have arose from this project.
-Outside of exploring how the performance of neural networks will vary with forecast horizon, we think that some other promising directions for exploration include cross-learning and hybrid models.
-Throughout this study, we have trained models individually for each target time series; cross-learning presents a different approach where we would train one model across all the time series available for a given target variable.
-Through cross-learning, the model might be able to learn a pattern from one time series that will improve forecasts on a different time series; but, if the model is trained on a single time series, then the model may never learn this pattern.
+In place of the one question that we set out to answer, namely, "How do neural network models perform for water quality forecasting?", we have unearthed an abundance of more questions and ideas for future work.
+Outside of exploring how the performance of neural networks will vary with forecast horizon, some other promising directions for exploration include cross-learning and hybrid models.
+Throughout this study, we have trained individual models for each target time series.
+Cross-learning presents a different approach where we would train one model across all the time series available for a given target variable.
+Through cross-learning, the model might be able to learn a pattern from one time series that will transfer to making accurate forecasts on a different time series.
 We also reckon that hybrid models may have superior forecasting performance to neural network or statistical models individually.
@@
 It is an exciting time to study water quality forecasting.
-There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as more sensors will be deployed \cite{}.
-Neural network models offer promising advantages over alternative methods for limnological forecasting as these methods can learn complex patterns without being restrained by a need for domain knowledge.
+There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as the cost for sensors and remote sensing continue to come down.
+Neural network models offer promising advantages over alternative methods for limnological forecasting as these methods can learn complex patterns from data without being restrained by a need for domain knowledge.
 Furthermore, their performance has the potential to improve as more water quality data becomes available.
 Yet, they also present a range of problems like a lack of interpretability and generalizability that could restrain them from being useful decision support tools for the safety critical problem of water resource management.
-This study does not attempt to provide a definitive answer as to whether neural networks will be the best method for water quality forecasting as answering this questions will take numerous studies to reach a consensus on.
-Instead, we hope that this study will generate ideas and a collective momentum towards improving our general ability to forecast and manage water quality.
+This study does not attempt to provide a definitive answer as to whether neural networks will be the best method class moving forward as answering this question will take a plethora of research.
+Instead, we hope that this study will generate more ideas and incite momentum towards improving our ability to forecast and manage water quality.
 

### 2024-06-27 manuscript/manuscript.Rmd
RStudio freezing on me

@@
 
+\begin{table}[!htbp]
+    \centering
+    \begin{subtable}{\textwidth}
+        \centering
+        \begin{tabular}{lcc}
+            \toprule
+            Abbreviation & Algorithm Name \\
+            \midrule
+            RNN & Recurrent Neural Network \cite{hochreiter_long_1997}  \\
+            BlockRNN & Block Recurrent Neural Network \cite{du_multivariate_2020} \\
+            TCN & Temporal Convolutional Network \cite{bai_empirical_2018} \\
+            NLinear & NLinear \cite{zeng_are_2022}  \\
+            DLinear &  DLinear \cite{zeng_are_2022} \\
+            TFT & Temporal Fusion Transformer \cite{lim_temporal_2019} \\
+            Transformer & Transformer \cite{vaswani_attention_2017} \\
+        \end{tabular}
+    \end{subtable}
+    \caption{The neural network-based time series forecasting models used in this study.}
+    \label{tab:ml_models}
+\end{table}
+
 # Results
@@
 
-In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables.
+In Table \ref{tab:mean_crps_rmse}, we present the mean CRPS and RMSE scores for the respective target variables.
 For dissolved oxygen (DO) and water temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, naive persistence and climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE.
@@
 
-By examining some of the individual forecasts, as shown in Figure \ref{}, it is possible to gain intuition for why the neural network models perform well across the target variables.
+By examining some of the individual forecasts, as shown in Figure \ref{fig:individual_forecasts}, it is possible to gain intuition for why the neural network models perform well across the target variables.
 The AutoTheta model produces forecasts that resemble linear regressions based upon recently observed values.
@@
 
-These general patterns in model performance can be observed in the skill score plots displayed in Figures \ref{}.
+These general patterns in model performance can be observed in the skill score plots displayed in Figures \ref{fig:oxygen_skill}, \ref{fig:temperature_skill}, \ref{fig:chla_skill}.
 The AutoTheta model tends to have longer tails, indicating that the AutoTheta forecasts are more likely to perform very well or very badly.
@@
 When examining the performance of the models within a forecast horizon, additional nuances emerge, confusing the perspective that neural network models are the best performing model class.
-In Figure \ref{}, the AutoTheta model is consistently the best performing model for short time horizons across target variables, but by the end of the 30-day horizon, the AutoTheta model is the worst perfoming model universally.
+In Figure \ref{fig:intra_skill}, the AutoTheta model is consistently the best performing model for short time horizons across target variables, but by the end of the 30-day horizon, the AutoTheta model is the worst perfoming model universally.
 Meanwhile the neural network models underperform AutoTheta during the early stages of the forecast window (generally, when t < 5), but their forecast skill declines less rapidly than the skill score of AutoTheta.
@@
 The neural network models tend to have shorter tails in the underperforming region.}
+  \label{fig:oxygen_skill}
 \end{figure}
@@
 This indicates that the neural network models are producing high accuracy forecasts more frequently than the AutoTheta model.}
+  \label{fig:temperature_skill}
 \end{figure}
@@
 The naive persistence null model attained the lowest RMSE on chlorophyll-a, and this is apparent in the RMSE-SS plot as the models have skill score distributions that are generally centered above the underperformance threshold.}
+  \label{fig:chla_skill}
 \end{figure}
@@
 The neural network models do not perform as well as AutoTheta at the beginning of the forecast horizon, but their performance declines less rapidly throughout the horizon.}
+  \label{fig:intra_skill}
 \end{figure}
@@
 With chlorophyll-a, the neural networks take a conservative strategy where they predict that chlorophyll-a will remain in a non-bloom state, and these models never take a risk in forecasting that a bloom will occur.}
+  \label{fig:individual_forecasts}
 \end{figure}
@@
 The neural network models struggled to forecast chlorophyll-a, a result that is not surprising given that forecasting algae blooms is well established as a challenging prediction problem in limnology \cite{chen_combining}.
-With cholorophyll-a, the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state.
+With chlorophyll-a, the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state.
 It is possible that the neural network models settled on this conservative behavior because we did provide enough information to predict blooms; the neural networks were not provided with some of the typical covariates used in process-based models like Nitrogen and Phosphorous concentrations, and photosynthetic active radiation.
@@
 
-In place of the one question that we set out to answer, namely, "How do neural network models perform for water quality forecasting?", we have unearthed an abundance of more questions and ideas for future work.
-Outside of exploring how the performance of neural networks will vary with forecast horizon, some other promising directions for exploration include cross-learning and hybrid models.
-Throughout this study, we have trained individual models for each target time series.
-Cross-learning presents a different approach where we would train one model across all the time series available for a given target variable.
-Through cross-learning, the model might be able to learn a pattern from one time series that will transfer to making accurate forecasts on a different time series.
-We also reckon that hybrid models may have superior forecasting performance to neural network or statistical models individually.
-We observed that AutoTheta were the best performing models early in the forecast horizon, and neural networks performed best later in the forecast horizon.
+This study had the simple motivation of comducting a model comparison, and has resulted in an abundance of research questions and ideas for future work.
+Outside of exploring how the performance of neural networks will vary with forecast horizon, other promising directions for exploration include cross-learning and hybrid models.
+Throughout this study, we trained individual models for each target time series.
+Cross-learning presents a different approach where we would train one model across multiple time series (potentially, all the time series for a give target variable or even across target variables).
+Through cross-learning, the model might be able to learn patterns from one time series that will transfer to making accurate forecasts on a different time series.
+Along these lines, developing a model hybridized across modeling classes could yield superior performance.
+Since we observed that AutoTheta was the best performing model early in the forecast horizon, and neural networks were the best performers later in the forecast horizon, it is sensible that a hybrid ML-statistical model could achieve better performance than a purely statistical or machine learning model alone.
+For instance, one could explore a model that heavily relies on a statistical model the beginning of a forecast horizon then favors a ML model towards the  when  longer horizons.
 This motivates the exploration of a hybrid model that employs AutoTheta at the beginning of the forecast horizon and neural network models later in the forecast horizon.

### 2024-06-28 manuscript/manuscript.Rmd
End of day edits

@@
 
-This study had the simple motivation of comducting a model comparison, and has resulted in an abundance of research questions and ideas for future work.
+This study had the sole motivation of conducting a model comparison, and has resulted in an abundance of research questions and ideas for future work.
 Outside of exploring how the performance of neural networks will vary with forecast horizon, other promising directions for exploration include cross-learning and hybrid models.
 Throughout this study, we trained individual models for each target time series.
-Cross-learning presents a different approach where we would train one model across multiple time series (potentially, all the time series for a give target variable or even across target variables).
-Through cross-learning, the model might be able to learn patterns from one time series that will transfer to making accurate forecasts on a different time series.
-Along these lines, developing a model hybridized across modeling classes could yield superior performance.
-Since we observed that AutoTheta was the best performing model early in the forecast horizon, and neural networks were the best performers later in the forecast horizon, it is sensible that a hybrid ML-statistical model could achieve better performance than a purely statistical or machine learning model alone.
-For instance, one could explore a model that heavily relies on a statistical model the beginning of a forecast horizon then favors a ML model towards the  when  longer horizons.
-This motivates the exploration of a hybrid model that employs AutoTheta at the beginning of the forecast horizon and neural network models later in the forecast horizon.
+Cross-learning presents a different approach where one model would be trained across multiple time series.
+Recently, cross-learning has been shown to improved forecast performance as models that employ cross-learning are able to learn patterns across different time series that transfer to making more accurate forecasts on an individual time series, particularly when there is a limited amount of data points in the individual time series  \cite{time_series}.
+Along these lines, developing a model hybridized across modeling classes could also yield superior performance.
+Since we observed that AutoTheta was the best performing model early in the forecast horizon, and neural networks were the best performers later in the forecast horizon, it is sensible that a hybrid ML-statistical model could achieve better performance than a purely statistical or machine learning model.
+For instance, a model that relies on a statistical model the beginning of a forecast horizon then favors a ML model towards the end of forecast window could achieve better performance than the models explored in this study.
 
 It is an exciting time to study water quality forecasting.
-There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as the cost for sensors and remote sensing continue to come down.
+There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as the cost for sensors and remote sensing continue to decline.
 Neural network models offer promising advantages over alternative methods for limnological forecasting as these methods can learn complex patterns from data without being restrained by a need for domain knowledge.
-Furthermore, their performance has the potential to improve as more water quality data becomes available.
-Yet, they also present a range of problems like a lack of interpretability and generalizability that could restrain them from being useful decision support tools for the safety critical problem of water resource management.
-This study does not attempt to provide a definitive answer as to whether neural networks will be the best method class moving forward as answering this question will take a plethora of research.
-Instead, we hope that this study will generate more ideas and incite momentum towards improving our ability to forecast and manage water quality.
+Furthermore, the performance of neural networks has the potential to improve as more water quality data becomes available.
+Yet, neural networks also present a range of problems like a lack of interpretability and generalizability that could restrain them from being useful decision support tools for the safety critical problem of water resource management.
+This study does not attempt to provide a definitive answer as to whether neural networks will be the best method class moving forward as resolving such a question will require a much larger body of work than what is presented here.
+Instead, the hope is that this study will generate ideas and incite momentum towards improving our ability to forecast and manage water quality.
 
